---
title: "Processing and analysis of MALDI-imaging data"
date: "`r BiocStyle::doc_date()`"
package: "`r BiocStyle::pkg_ver('gutenTAG')`"
author:
- name: John Abbey
  affiliation: 
  - Department for Quantitative Biomedicine, University of Zurich
  - Institute for Molecular Health Sciences, ETH Zurich
  - Functional Genomics Center Zurich
  email: john.abbey@uzh.ch
  
- name: Pierre Bost
  affiliation:
  - Department for Quantitative Biomedicine, University of Zurich

- name: Leonardo Schwarz
  affiliation:
    - Functional Genomics Center Zurich
    
output:
    BiocStyle::html_document:
        toc_float: yes
        
bibliography: library.bib

abstract: |
    This R package supports the processing and analysis of highly multiplexed MALDI-imaging data. `gutenTAG` is aimed at processing of highly-multiplexed targeted MALDI-IHC data in particular. It is characterised by a novel 'metapeak' generation step to account for mass shift and the presence of isotopic peaks. The data handling capabilities build heavily on the R package `Cardinal`, a very useful tool for general processing and analysis of Mass Spectrometry Imaging data. The data can easily be converted into `MSImagingExperiment`, `SpatialExperiment` objects and `AnnData` objects through inbuilt conversion functions, facilitating further downstream analysis. Additionally, the package supports image visualisation from `cytomapper` (and therefore `cytoviewer`, an RShiny extension of cytomapper) through the construction of a `CytoImageList` object.
    
vignette: |
    %\VignetteIndexEntry{"Tools for MALDI-imaging data analysis"}
    %\VignetteEncoding{UTF-8}
    %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, echo=FALSE, results="hide"}
knitr::opts_chunk$set(error=FALSE, warning=FALSE, message=FALSE, crop = NULL)
#library(BiocStyle)
```

```{r library, echo=FALSE}
library(gutenTAG)
```

# Introduction

This vignette introduces the `gutenTAG` package for processing
highly-multiplexed MALDI-IHC images. `gutenTAG` is built heavily on
`r Biocpkg("Cardinal")` Bioconductor package [@bemis2023cardinal], a
go-to resource for processing and analysis of Mass Spectrometry Imaging
data. `gutenTAG` uses relevant functionality from `Cardinal` as well as
a novel 'metapeak' generation step to facilitate accurate peak picking
in the presence of technical mass shift and isotopic peaks. \`gutenTAG\`
is designed specifically for cases such as MALDI-IHC with photocleavable
mass tags (PCMTs), where we are searching for known features that are
spiked into our samples.

## Highly-multiplexed imaging

The development of highly-multiplexed imaging technologies has enabled
the exploration and characterisation of tissues on a spatial level
hitherto unimaginable. Advances in these technologies have enabled
analysis of the complex systems at play in diseases such as cancer in a
manner that respects the spatial nature of their progression. Spatial
transcriptomic methods have dominated the spatial biology landscape in
recent years. This is in large part due to the massive feature spaces
that emerge from these technologies, opening the door to the application
of high-dimensional statistics and machine learning techniques on
molecular biology datasets. While spatial transcriptomics has captured
the limelight, there is a large and ever-growing interest in
highly-multiplexed proteomics imaging due to the foundational role that
proteins play in cellular processes and in disease [@Moffitt2022]. This
field has seen a huge amount of innovation in recent years, resulting in
the emergence of myriad different approaches to achieve
highly-multiplexed targeted proteomic imaging. These can be broadly
characterised into cyclic immunofluorescence optical approaches, such as
CyCIF [@Lin2018], and one-step mass tag methods such as IMC
[@Giesen2014]. However, the high-plexities achieved by methods in
spatial transcriptomics has eluded the field of spatial proteomics so
far.

## Mass Spectrometry Imaging

Mass Spectrometry Imaging (MSI), a powerful imaging technology with many
applications in metabolomics, lipidomics, glycomics, etc, has largely
been excluded from the proteomics party. Proteins are typically too
large for direct analysis by mass spectrometry imaging. MSI methods have
been developing at a rapid rate in recent years. One such method is the
introduction of Photocleavable Mass Tags (PCMTs) [@ambergen], which
enable targeted tagging of proteins of interest via specific antibodies.
This technique of applying PCMT-tagged antibodies with MALDI-imaging has
been termed MALDI-IHC. The use of peptide tags as the mass reporter
system of choice lifts the lid on the existing plexity ceiling in
current one-shot imaging proteomic technologies, as it is no longer
capped by the abundance of heavy metals as in IMC.

## MALDI-IHC

MALDI-IHC data shows some persistent features which must be taken into
consideration for analysis. Firstly, it is large. Very large. Depending
on the size of the tissue in question and the pixel size chosen for
acquisition, it is not uncommon to see files in excess of 50GB. This
makes running even computationally efficient tasks quite a burden. On
the technical side, MALDI-IHC, and indeed MALDI-imaging in general,
commonly exhibit a significant mass shift. This can hamper our analysis,
as our peaks are not always in their expected location. This mass shift
can vary from sample to sample as well as vary within the same sample
(can be due to height effects in the tissue, laser defocusing, etc.). A
further feature of MALDI-imaging data is the presence of isotopic peaks,
which split the signal locally across the spectrum. These are the
primary features of MALDI-imaging data which can be corrected for
computationally using `gutenTAG`.

## gutenTAG overview

`gutenTAG` is an R package which forms a pipeline that enables fast,
robust and accurate processing of targeted MALDI-imaging data. It is
initiated by [reading](#read-in-data) .imzML data into R using the
`Cardinal` `readImzML` or `readMSIData` functions. It is then
recommended to [pre-process](#preprocessing) the data. This is followed
by [peak picking](#peak-detection) to identify peaks filter out noise in
our spectra. Following this is the novel
[metapeak](#metapeak-generation) step, which identifies the *m/z*
location of our peaks based on watershed-like algorithm performed on the
counts histogram of peaks. These locations on the *m/z* axis are called
metapeaks since they reflect the prominent location of our peaks across
all spectra in our image. Finally, these metapeaks are then mapped onto
our reference list of expected mass tags (our panel) and our metapeaks
are [assigned](#metapeak-assignment) a label from our panel, completing
the pipeline. This data is provided in a simple list format rather than
the typical object-oriented framework you may be familiar with from many
packages. Functions for easily generating many of the common object
classes in the spatial biology field are provided. The object classes
supported by `gutenTAG` include the `MSImagingExperiment` object from
`Cardinal`, the `CytoImageList` object for `cytomapper` and `cytoviewer`
and the popular `SpatialExperiment` [@spe2022] and `AnnData` objects.
This is a design choice, since the MALDI-IHC project bridges people from
many communities who may be familiar with different frameworks.

# gutenTAG pipeline

The `gutenTAG` package exports functions and example data to perform the
following analyses:

1.  [Read in raw datadata](#read-in-data)
2.  [Perform data pre-processing](#preprocessing)
3.  [Data processing](#processing)
4.  [Visualisation](#visualisation) of processed targeted images
5.  [Interoperability](#interoperability) with object classes

To highlight the usability of these function, the `gutenTAG` package
also exports a number of [test data files](#test-data).

## Read in raw data and panel {#read-in-data}

The raw imaging data can be read into R using the `readMSIData` function
from Cardinal.

```{r read in data}
imzML_path <- system.file("extdata/Example_data.imzML", package = "gutenTAG")
rawFile <- Cardinal::readMSIData(imzML_path)
```

The `readPanel` function can be used to read a .csv file into R.
containing the information for the mass of the peptide tag and the
molecular target into R.

```{r read in panel}
panel_path <- system.file("extdata/ref_list.csv", package = "gutenTAG")
panel <- readPanel(panel_path)
```

### Example data {#test-data}

If you want to see how the package functions on test data, or do not
have access to the required data yet, a good place to start is by
loading in the example data from the github repository. The `gutenTAG`
package contains example data generated by Bruker's RapiFlex imaging
system. The following section gives an overview of these files.

These files are accessible via:

```{r example data}
path <- system.file("extdata", package = "gutenTAG")
list.files(path, recursive = TRUE)
```

By default, the imaging data is read into an
`r BiocStyle::Biocpkg("MSImagingExperiment")` object. Here, the
extracted channel-specific intensities are stored in the `imageData`
slot, accessible via the `iData` getter function from `Cardinal`.

## Pre-processing {#preprocessing}

When acquiring the raw data, the resulting spectra can often be noisy
and messy. To correct for this, some pre-processing steps are required.
This can be done simply using the `preProcess` function from this
package, which can be implemented as follows.

```{r pre-processing}
pre <- preProcess(rawFile, cores = 2)
```

This function is a wrapper for the Cardinal functions `normalize`,
`smoothSignal` and `reduceBaseline`, using common parameters and methods
for each function. For more flexibility in your pre-processing, it is
recommended to directly use the Cardinal functions. Here, we use total
ion current (TIC) as the normalisation method of choice, Gaussian
smoothing for spectral smoothing and local minimum baseline reduction.

```{r Cardinal pre-processing}
pre <- rawFile |>
    Cardinal::normalize(method = "tic") |>
    smoothSignal(method = "gaussian", plot = FALSE) |>
    reduceBaseline(method = "locmin") |>
    process(BPPARAM = MulticoreParam(workers = 2))

```

## Processing {#processing}

Full processing of the pre-processed MALDI-imaging data can be performed
using the `peakDetection`, `generateMetapeaks` and `assignMetapeaks`
functions in series.

### Peak Detection {#peak-detection}

The first step of processing is to identify peaks in the pre-processed
spectra. This function is essentially a wrapper for
`Cardinal::peakPick`, and uses the median adaptive deviation (MAD) peak
detection algorithm for robust noise evaluation.

```{r Peak detection}
detected <- peakDetection(pre, snr = 3, cores = 2)
```

### Metapeak generation {#metapeak-generation}

Once `peakDetection` has been performed, the detected peaks per pixel
can be used to generate metapeaks. Here, metapeaks are defined as single
peaks that correspond to clusters of peaks that contain information on
the same molecular species. The function used to generate these
metapeaks is `generateMetapeaks`. This step is motivated by the loss of
signal through technical mass shifts as well as the presence of isotopic
peaks in our spectra. The following steps are required to be construct
metapeaks.

1.  Obtain the number of pixels in which each detected peak is present
    (counts histogram).
2.  Gaussian smoothing over the counts histogram.
3.  Watershed algorithm to cluster detected peaks
4.  Filter peaks present in less than a given threshold of pixels, which
    can be controlled using the `threshold` parameter.

It is important to note that metapeaks are not restricted to targeted
peaks. Many of the metapeaks generated will be formed by signal that is
from untargeted peaks.

```{r Metapeak generation}
metapeaks <- generateMetapeaks(detected, hist_smooth_factor = 1, fixed.limits = 0.5, sparsity = 1)
```

### Assigning metapeaks {#metapeak-assignment}

The final step of processing is to assign metapeaks to the reference
list of masses that was loaded in using the `readPanel` function. This
is performed using the `assignMetapeaks` function. The output of this
function is a list of lists, containing primarily `IntensityDF`, the
channel-specific intensities, `SpatialCoords`, the spatial coordinates
and `CorrespondenceMatrix`, a data frame describing the assignment of
metapeaks to reference peaks. This assignment can be controlled via the
`mz_threshold` parameter, which has a default of 1*m/z*.

You may notice that some of the expected peaks in the reference list
will not appear in the output. This is likely due to either a lack of
signal, resulting in the metapeak not passing the filter threshold
during `generateMetapeaks`, or the metapeak center being out of range
for assignment to a reference peak during `assignMetapeaks`. These
parameters can be adjusted to improve results.

```{r Generate final output}
processed <- assignMetapeaks(metapeaks, refList = panel, pre = pre, mz_threshold = 1)
```

## Visualisation of processed data {#visualisation}

### Cytomapper

This package supports visualisation of MALDI images through the
`cytomapper` [@Eling2020] framework developed by Eling et al
(reference). In order to use the `plotPixels` function, the data must
first be coerced to a `CytoImageList` object using the `asCytoImageList`
function from this package.

```{r coerce to CytoImageList object}

library(cytomapper)
cil <- asCytoImageList(processed)

```

Now the channels can be visualised using the `plotPixels` function.
Single channel and multichannel imaging is supported using this format.

```{r single channel imaging}
cytomapper::plotPixels(image = cil, 
                       colour_by = "HLA.ABC", 
                       scale_bar = list(length = 5, lwidth = 0.5, margin = c(1,1)),
                       interpolate = FALSE, 
                       image_title = list(margin = c(1, 1)))
```

```{r multi-channel imaging}
plotPixels(cil, colour_by = c("HLA.ABC", "CD98", "beta.actin"),
           legend = list(colour_by.title.cex = 0.7), 
           scale_bar = list(length = 5, lwidth = 0.5, margin = c(1,1)),
           image_title = list(margin = c(1, 1)),
           interpolate = FALSE)
```

### Cytoviewer

Additionally, the images can be visualised using the `cytoviewer` R
Shiny app developed by Meyer et al 2023 [@Meyer2023]. `cytoviewer`
allows interactive visualistion of multi-channel images, supporting
visualisation of up to 6 channels. It should be noted that the
cell-level functionality of `cytomapper` and `cytoviewer` are not
applicable for our data, since single-cell resolution is not yet
possible. Interactive visualisation of our images can't be performed
through a vignette, but the call is very simple: `cytoviewer(cil)`.

## Interoperability {#interoperability}

`gutenTAG` provides functions for easy conversion to some of the popular
frameworks in the spatial omics ecosystem. These functions include:

-   `asMSImagingExperiment()` for `MSImagingExperiment` class from
    `Cardinal` for those who are familiar with typical MSI analysis and
    prefer to return to the `Cardinal` framework

-   `asSpatialExperiment()` for `SpatialExperiment` class, a spatial
    extension of the `SingleCellExperiment` class.

-   `asCytoImageList()` for `CytoImageList` class for image
    visualisation using the `cytomapper` and `cytoviewer` packages.

-   `asAnnData()` for `AnnData` class for access to the `scverse`
    ecosystem.

These functions are written to be called on the output of the
`assignMetapeaks()` function.

# Contributions

[Pierre Bost](https://github.com/PierreBSC) conceptualised the metapeak
generation step for data processing, as well as being instrumental in
developing the architecture of the pipeline.\ [Leonardo Schwarz](https://github.com/leoschwarz) was contributed enormously  
to the package codebase.\ Thank you to [Nils Eling](https://github.com/nilseling) and [Lasse
Meyer](https://github.com/lassedochreden) for assisting in the development of the package, as well as helping with 
integrating the package with `cytomapper` and `cytoviewer`. \ 
We thank all the members of FGCZ Proteomics Informatics unit for their support.

# Session info {.unnumbered}

```{r sessionInfo, echo=FALSE}
sessionInfo()
```

# References
