---
title: "Processing and analysis of MALDI-imaging data"
date: "`r BiocStyle::doc_date()`"
package: "`r BiocStyle::pkg_ver('gutenTAG')`"
author:
- name: John Abbey
  affiliation: 
  - Department for Quantitative Biomedicine, University of Zurich
  - Institute for Molecular Health Sciences, ETH Zurich
  - Functional Genomics Center Zurich
  email: john.abbey@uzh.ch
  
- name: Pierre Bost
  affiliation:
  - Department for Quantitative Biomedicine, University of Zurich

- name: Leonardo Schwarz
  affiliation:
    - Functional Genomics Center Zurich
    
output:
    BiocStyle::html_document:
        toc_float: yes
        
bibliography: library.bib

abstract: |
    This R package supports the processing and analysis of highly multiplexed MALDI-imaging data. `gutenTAG` is aimed at processing of highly-multiplexed targeted MALDI-IHC data in particular. It is characterised by a novel 'metapeak' generation step to account for mass shift and the presence of isotopic peaks. The data handling capabilities build heavily on the R package Cardinal, a very useful tool for general process and analysis of Mass Spectrometry Imaging data. The data can easily be converted into `MSImagingExperiment`, `SpatialExperiment` [@spe2022] objects and `AnnData` objects through inbuilt conversion functions, facilitating further downstream analysis. Additionally, the package supports image visualisation from `cytomapper` (and therefore `cytoviewer`, an RShiny extension of cytomapper) through the construction of a `CytoImageList` object.
    
vignette: |
    %\VignetteIndexEntry{"Tools for MALDI-imaging data analysis"}
    %\VignetteEncoding{UTF-8}
    %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, echo=FALSE, results="hide"}
knitr::opts_chunk$set(error=FALSE, warning=FALSE, message=FALSE, crop = NULL)
#library(BiocStyle)
```

```{r library, echo=FALSE}
library(gutenTAG)
```

# Introduction

`r Biocpkg("Cardinal")`

Need to write introduction.

## Overview

The `gutenTAG` package exports functions and example data to perform
the following analyses:

1.  [Read in raw datadata](#read-in-data)
2.  [Perform data pre-processing](#preprocessing)
3.  [Data processing](#processing)
4.  [Visualisation](#visualisation) of processed targeted images


To highlight the usability of these function, the `gutenTAG` package
also exports a number of [test data files](#test-data).

# Read in raw data and panel {#read-in-data}

The raw imaging data can be read into R using the `readMSIData` function
from Cardinal [@bemis2023cardinal].

```{r read in data}
imzML_path <- system.file("extdata/Example_data.imzML", package = "gutenTAG")
rawFile <- Cardinal::readMSIData(imzML_path)
```

The `readPanel` function can be used to read a .csv file into R.
containing the information for the mass of the peptide tag and the
molecular target into R.

```{r read in panel}
panel_path <- system.file("extdata/ref_list.csv", package = "gutenTAG")
panel <- readPanel(panel_path)
```

# Example data {#test-data}

If you want to see how the package functions on test data, or do not
have access to the required data yet, a good place to start is by
loading in the example data from the github repository. The
`gutenTAG` package contains example data generated by Bruker's RapiFlex
imaging system. The following section gives an overview of these files.

These files are accessible via:

```{r example data}
path <- system.file("extdata", package = "gutenTAG")
list.files(path, recursive = TRUE)
```

By default, the imaging data is read into a
`r BiocStyle::Biocpkg("MSImagingExperiment")` object. Here, the
extracted channel-specific intensities are stored in the `???` slot.

# Data pre-processing {#preprocessing}

When acquiring the raw data, the resulting spectra can often be noisy
and messy. To correct for this, some pre-processing steps are required.
This can be done simply using the `preProcess` function from this
package, which can be implemented as follows.

```{r pre-processing}
pre <- preProcess(rawFile, cores = 2)
```

This function is a wrapper for the Cardinal functions `normalize`,
`smoothSignal` and `reduceBaseline`, using common parameters and methods
for each function. For more flexibility in your pre-processing, it is
recommended to directly use the Cardinal functions. Here, we use total
ion current (TIC) as the normalisation method of choice, Gaussian
smoothing for spectral smoothing and local minimum baseline reduction.

```{r Cardinal pre-processing}
pre <- rawFile %>%
    Cardinal::normalize(method = "tic") %>%
    smoothSignal(method = "gaussian", plot = FALSE) %>%
    reduceBaseline(method = "locmin") %>%
    process(BPPARAM = MulticoreParam(workers = 2))

```

# Data processing {#processing}

Full processing of the pre-processed MALDI-imaging data can be performed
using the `peakDetection`, `generateMetapeaks` and `assignMetapeaks`
functions in series.

## Peak Detection

The first step of processing is to identify peaks in the pre-processed
spectra. This function is essentially a wrapper for
`Cardinal::peakPick`, and uses the median adaptive deviation (MAD) peak
detection algorithm for robust noise evaluation.

```{r Peak detection}
detected <- peakDetection(pre, snr = 3, cores = 2)
```

## Metapeak generation

Once `peakDetection` has been performed, the detected peaks per pixel
can be used to generate metapeaks. Here, metapeaks are defined as single
peaks that correspond to clusters of peaks that contain information on
the same molecular species. The function used to generate these
metapeaks is `generateMetapeaks`. This step is motivated by the loss of signal through technical mass shifts as well as the presence of isotopic peaks in our spectra. The following steps are required to be construct metapeaks.

1.  Obtain the number of pixels in which each detected peak is
    present (counts histogram).
2.  Gaussian smoothing over the counts histogram.
3.  Watershed algorithm to cluster detected peaks
4.  Filter peaks present in less than a given threshold of pixels, which
    can be controlled using the `threshold` parameter.

It is important to note that metapeaks are not restricted to targeted
peaks. Many of the metapeaks generated will be formed by signal that is
from untargeted peaks.

```{r Metapeak generation}
metapeaks <- generateMetapeaks(detected)
```

## Assigning metapeaks

The final step of processing is to assign metapeaks to the reference
list of masses that was loaded in using the `readPanel` function. This
is performed using the `assignMetapeaks` function. The output of this
function is a list of lists, containing primarily `IntensityDF`, the
channel-specific intensities, `SpatialCoords`, the spatial coordinates
and `CorrespondenceMatrix`, a data frame describing the assignment of
metapeaks to reference peaks. This assignment can be controlled via the
`mz_threshold` parameter, which has a default of 1m/z.

You may notice that some of the expected peaks in the reference list
will not appear in the output. This is likely due to either a lack of
signal, resulting in the metapeak not passing the filter threshold
during `generateMetapeaks`, or the metapeak center being out of range
for assignment to a reference peak during `assignMetapeaks`. These
parameters can be adjusted to improve results.

```{r Generate final output}
processed <- assignMetapeaks(metapeaks, refList = panel, pre = pre)
```

# Visualisation of processed data {#visualisation}

## Cytomapper

This package supports visualisation of MALDI images through the
`cytomapper` [@Eling2020]  framework developed by Eling et al (reference). In order to
use the `plotPixels` function, the data must first be coerced to a
`CytoImageList` object using the `asCytoImageList` function from this
package.

```{r coerce to CytoImageList object}
library(cytomapper)
cil <- asCytoImageList(processed)

```

Now the channels can be visualised using the `plotPixels` function.
Single channel and multichannel imaging is supported using this format.

```{r single channel imaging}
cytomapper::plotPixels(image = cil, 
                       colour_by = "HLA.ABC", 
                       scale_bar = list(length = 5, lwidth = 0.5, margin = c(1,1)),
                       interpolate = FALSE, 
                       image_title = list(margin = c(1, 1)))
```

```{r multi-channel imaging}
plotPixels(cil, colour_by = c("HLA.ABC", "CD98", "beta.actin"),
           legend = list(colour_by.title.cex = 0.7), 
           scale_bar = list(length = 5, lwidth = 0.5, margin = c(1,1)),
           image_title = list(margin = c(1, 1)),
           interpolate = FALSE)
```

## Cytoviewer

Additionally, the images can be visualised using the `cytoviewer` R
Shiny app developed by Meyer et al 2023 [@Meyer2023]. `cytoviewer` allows
interactive visualistion of multi-channel images, supporting
visualisation of up to 6 channels. It should be noted that the
cell-level functionality of `cytomapper` and `cytoviewer` are not
applicable for our data, since single-cell resolution is not yet
possible. Interactive visualisation of our images can't be performed
through a vignette, but the call is very simple: `cytoviewer(cil)`.


# Contributions

[Pierre Bost](https://github.com/PierreBSC) conceptualised the metapeak
generation step for data processing, as well as being instrumental in
developing the architecture of the pipeline.\ [Nils Eling](https://github.com/nilseling) assisted in designing and
developing the structure of the package, as well as assisting in
integrating the package with `cytomapper`, while [Lasse
Meyer](https://github.com/lassedochreden) has contributed to integrating
the package with `cytoviewer`.

# Session info {.unnumbered}

```{r sessionInfo, echo=FALSE}
sessionInfo()
```

# References
