---
title: "Processing and analysis of MALDI-imaging data"
date: "`r BiocStyle::doc_date()`"
package: "`r BiocStyle::pkg_ver('maldipackage')`"
author:
- name: John Abbey
  affiliation: 
  - Department for Quantitative Biomedicine, University of Zurich
  - Institute for Molecular Health Sciences, ETH Zurich
  email: john.abbey@uzh.ch
output:
    BiocStyle::html_document:
        toc_float: yes
#bibliography: library.bib
abstract: |
    This R package supports the processing and analysis of highly multiplexed MALDI-imaging data. The data handling capabilities (import, pre-processing, export) build heavily on the R package Cardinal, designed for general Mass Spectrometry Imaging data. The main functionality of this package includes reliable data processing, extensive quality control and data analysis. First, data is processing using a novel approach involving the construction of metapeaks to account for mass shift and the presence of isotopic peaks. Then image-level and spatial quality control statistics, such as computing Geary's C scores can be computed for every marker. Furthermore, spatial analysis can be performed on the high-dimensional images, including constructing variograms and a variety of clustering algorithms. The package also supports dimensionality reduction methods, including PCA, NMF and UMAP. Additionally, the package supports image visualisation from cytomapper (and cytoviewer) through the construction of a CytoImageList object. As such, once the data is coerced to this format, many of the analysis techniques that are applicable to CytoImageList can be applied.
    
vignette: |
    %\VignetteIndexEntry{"Tools for IMC data analysis"}
    %\VignetteEncoding{UTF-8}
    %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, echo=FALSE, results="hide"}
knitr::opts_chunk$set(error=FALSE, warning=FALSE, message=FALSE, crop = NULL)
#library(BiocStyle)
```

```{r library, echo=FALSE}
library(maldipackage)
```

# Introduction

`r Biocpkg("Cardinal")`

## Overview

The `maldipackage` package exports functions and example data to perform
the following analyses:

1.  [Read in raw datadata](#read-in-data)
2.  [Perform data pre-processing](#preprocessing)
3.  [Data processing](#processing)
4.  [Visualisation](#visualisation) of processed targeted images
5.  [Dimensionality reduction](#dimensionality-reduction) of processed
    targeted images
6.  [Community detection](#community-detection)

To highlight the usability of these function, the `maldipackage` package
also exports a number of [test data files](#test-data).

# Read in raw data and panel {#read-in-data}

The raw imaging data can be read into R using the `readMSIData` function
from Cardinal.

```{r read in data}
imzML_path <- system.file("extdata/Example_data.imzML", package = "maldipackage")
rawFile <- Cardinal::readMSIData(imzML_path)
```

The `readPanel` function can be used to read a .csv file into R.
containing the information for the mass of the peptide tag and the
molecular target into R.

```{r read in panel}
panel_path <- system.file("extdata/ref_list.csv", package = "maldipackage")
panel <- readPanel(panel_path)
```

# Example data {#test-data}

If you want to see how the package functions on test data, or do not
have access to the required data yet, a good place to start is by
loading in the example data from the github repository. The
`maldipackage` package contains example data generated by the RapiFlex
imaging system. The following section gives an overview of these files.

These files are accessible via:

```{r example data}
path <- system.file("extdata", package = "maldipackage")
list.files(path, recursive = TRUE)
```

By default, the imaging data is read into a
`r BiocStyle::Biocpkg("MSImagingExperiment")` object. Here, the
extracted channel-specific intensities are stored in the `???` slot.

# Data pre-processing {#preprocessing}

When acquiring the raw data, the resulting spectra can often be noisy
and messy. To correct for this, some pre-processing steps are required.
This can be done simply using the `preProcess` function from this
package, which can be implemented as follows.

```{r pre-processing}
pre <- preProcess(rawFile, cores = 2)
```

This function is a wrapper for the Cardinal functions `normalize`,
`smoothSignal` and `reduceBaseline`, using common parameters and methods
for each function. For more flexibility in your pre-processing, it is
recommended to directly use the Cardinal functions. Here, we use total
ion current (TIC) as the normalisation method of choice, Gaussian
smoothing for spectral smoothing and local minimum baseline reduction.

```{r Cardinal pre-processing}
pre <- rawFile %>%
    Cardinal::normalize(method = "tic") %>%
    smoothSignal(method = "gaussian", plot = FALSE) %>%
    reduceBaseline(method = "locmin") %>%
    process(BPPARAM = MulticoreParam(workers = 2))

```

# Data processing {#processing}

Full processing of the pre-processed MALDI-imaging data can be performed
using the `peakDetection`, `generateMetapeaks` and `assignMetapeaks`
functions in series.

## Peak Detection

The first step of processing is to identify peaks in the pre-processed
spectra. This function is essentially a wrapper for
`Cardinal::peakPick`, and uses the median adaptive deviation (MAD) peak
detection algorithm for robust noise evaluation.

```{r Peak detection}
detected <- peakDetection(pre, snr = 3, cores = 2)
```

## Metapeak generation

Once `peakDetection` has been performed, the detected peaks per pixel
can be used to generate metapeaks. Here, metapeaks are defined as single
peaks that correspond to clusters of peaks that contain information on
the same molecular species. The function used to generate these
metapeaks is `generateMetapeaks`. This step is implemented to account
for loss of signal through technical shift and isotopic peaks in the
spectra. Inside this function, metapeaks are constructed by

1.  Obtaining the number of pixels in which each detected peak is
    present.
2.  Gaussian smoothing over the spectrum
3.  Watershed algorithm to cluster detected peaks
4.  Filter peaks present in less than a given threshold of pixels, which
    can be controlled using the `threshold` parameter.

It is important to note that metapeaks are not restricted to targeted
peaks. Many of the metapeaks generated will be formed by signal that is
from untargeted peaks.

```{r Metapeak generation}
metapeaks <- generateMetapeaks(detected)
```

## Assigning metapeaks

The final step of processing is to assign metapeaks to the reference
list of masses that was loaded in using the `readPanel` function. This
is performed using the `assignMetapeaks` function. The output of this
function is a list of lists, containing primarily `IntensityDF`, the
channel-specific intensities, `SpatialCoords`, the spatial coordinates
and `CorrespondenceMatrix`, a data frame describing the assignment of
metapeaks to reference peaks. This assignment can be controlled via the
`mz_threshold` parameter, which has a default of 1m/z.

You may notice that some of the expected peaks in the reference list
will not appear in the output. This is likely due to either a lack of
signal, resulting in the metapeak not passing the filter threshold
during `generateMetapeaks`, or the metapeak center being out of range
for assignment to a reference peak during `assignMetapeaks`. These
parameters can be adjusted to improve results.

```{r Generate final output}
processed <- assignMetapeaks(metapeaks, refList = panel, pre = pre)
```

# Visualisation of processed data {#visualisation}

## Cytomapper

This package supports visualisation of MALDI images through the
`cytomapper` framework developed by Eling et al (reference). In order to
use the `plotPixels` function, the data must first be coerced to a
`CytoImageList` object using the `asCytoImageList` function from this
package.

```{r coerce to CytoImageList object}
library(cytomapper)
cil <- asCytoImageList(processed)

```

Now the channels can be visualised using the `plotPixels` function.
Single channel and multichannel imaging is supported using this format.

```{r single channel imaging}
cytomapper::plotPixels(image = cil, 
                       colour_by = "HLA.ABC", 
                       scale_bar = list(length = 5, lwidth = 0.5, margin = c(1,1)),
                       interpolate = FALSE, 
                       image_title = list(margin = c(1, 1)))
```

```{r multi-channel imaging}
plotPixels(cil, colour_by = c("HLA.ABC", "CD98", "beta.actin"),
           legend = list(colour_by.title.cex = 0.7), 
           scale_bar = list(length = 5, lwidth = 0.5, margin = c(1,1)),
           image_title = list(margin = c(1, 1)),
           interpolate = FALSE)
```

## Cytoviewer

Additionally, the images can be visualised using the `cytoviewer` R
Shiny app developed by Meyer et al 2023 (ref). `cytoviewer` allows
interactive visualistion of multi-channel images, supporting
visualisation of up to 6 channels. It should be noted that the
cell-level functionality of `cytomapper` and `cytoviewer` are not
applicable for our data, since single-cell resolution is not yet
possible. Interactive visualisation of our images can't be performed
through a vignette, but the call is very simple: `cytoviewer(cil)`.


# Dimensionality Reduction {#dimensionality-reduction}

## Principal Component Analysis (PCA)

Dimensionality reduction techniques can be helpful in revealing the
overall structure of the data, especially during the exploratory phase
of data analysis. This package explicitly supports PCA, NMF and UMAP.

PCA is a very commonly used techniques which linearly separates the data
according to the orthogonal dimensions with greatest variance. This can
be performed using the `computePCA` function. This function wraps the
`irlba` implementation of PCA, requiring a set seed. This takes the
targeted intensity dataframe as the main argument. The number of
components, aswell as some plotting functions can also be chosen as
arguments.

Once this is performed, the loadings for each component can be accessed
under `my_pca$rotation`.

```{r pca}
#my_pca <- computePCA(processed$IntensityDF, comp = 3, seed = 123, scree = TRUE)
```

Plot PCA

```{r plot pca}
#for (i in 1:ncol(my_pca$x)){
#  imageChannel(my_pca$x, coords = processed$SpatialCoords, channel_number = i, axes = FALSE, colna = "black")
#}
```

## Non-negative Matrix Factorisation (NMF)

In practice, NMF often performs better than PCA in terms of decomposing
the major dimensions of the data. NMF gives parts-based representation
of the data in the feature factor matrix, denoted here as 'w'. NMF can
be simply executed using the `computeNMF` function provided by this
package.

```{r nmf}
my_nmf <- computeNMF(processed$IntensityDF, comp = 3)

# visualise the factored dimensions
for (i in 1:ncol(my_nmf$w)){
  imageChannel(my_nmf$w, coords = processed$SpatialCoords, channel_number = i)
}
```

## Uniform Manifold Approximate Projection (UMAP)

UMAP is a non-linear dimensionality technique that embeds high
dimensional data into two dimensions, providing an overview of the
global structure of the data.

This package includes the `computeUMAP` function, which computes the
UMAP using the `uwot` package implementation.

```{r umap}
my_umap <- computeUMAP(processed$IntensityDF, seed = 3)
```

One of the ways in which UMAP can be used is to display single-channel
intensity information on top of the UMAP.

```{r visualise single channel intensities on umap}
# create data frame
umap_df <- cbind(my_umap, processed$IntensityDF)

# plot UMAP with channel intensity
library(ggplot2)
ggplot(umap_df) +
  geom_point(aes(x = UMAP1, y = UMAP2, color = HLA.ABC), alpha = 0.6, size = 1) +
  scale_color_viridis_c() +
  theme_minimal() +
  theme(legend.position = "right") +
  labs(title = "HLA.DR",
       x = "UMAP 1",
       y = "UMAP 2",
       color = "Intensity")
```

# Community detection {#community-detection}

The `maldipackage` exports the `louvainCluster` function to detect the
spatial community of each pixel. Here, each pixe is clustered based on
its interactions as defined by a spatial object graph.

In more detail, the spatial community detection procedure is as follows:

1.  Create an igraph object.

2.  Perform community detection using the louvain algorithm.

3.  Store the community IDs in a vector.

```{r community detection}
clusters <- louvainCluster(processed$IntensityDF, processed$SpatialCoords, k = 30)
```

One of the ways in which these clusters can be visualised is to display
the cluster memberships over the UMAP. This shows us how cluster
formation corresponds to the global structure of the data.

```{r visualise single channel intensities on spatial coords}
# create data frame
cluster_df <- cbind(my_umap, clusters)

# plot UMAP with channel intensity
ggplot(cluster_df, aes(x = UMAP1, y = UMAP2, color = clusters)) +
  geom_point(size = 1) +
  scale_color_discrete(name = "Cluster Membership") +
  theme_minimal() +
  labs(title = "Cluster Memberships on Spatial Coordinates",
       x = "x",
       y = "y")

```

Perhaps more informatively, the cluster memberships can be plotted onto
the spatial coordinates, revealing how the spatial context of the tissue
informs the cluster membership.

```{r visualise single channel intensities on spatial coordinates}
# create data frame
cluster_coords <- cbind(processed$SpatialCoords, clusters)

# ggplot spatial distribution of clusters
ggplot(cluster_coords, aes(x = x, y = y, color = clusters)) +
  geom_point(size = 5) +
  scale_color_discrete(name="Cluster Membership") +
  #scale_color_brewer(palette = "Set1", name = "Cluster Membership") + # alternative colour set for clustering
  theme_minimal() +
  labs(title = "Cluster Memberships on Spatial Coordinates",
       x = "X Coordinate",
       y = "Y Coordinate")
```

# Contributions

[Pierre Bost](https://github.com/PierreBSC) conceptualised the metapeak
generation step for data processing, as well as being instrumental in
developing the architecture of the pipeline.\ [Nils Eling](https://github.com/nilseling) assisted in designing and
developing the structure of the package, as well as assisting in
integrating the package with `cytomapper`, while [Lasse
Meyer](https://github.com/lassedochreden) has contributed to integrating
the package with `cytoviewer`.

# Session info {.unnumbered}

```{r sessionInfo, echo=FALSE}
sessionInfo()
```

# References
