---
title: "Quality Control Report"
params:
    param_load_dev_path: NA
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    toc-expand: 1
    toc-title: Contents
    toc-location: right
    code-fold: true
    code-tools: true
    code-summary: "Show the code"
    css: styles.css
    smooth-scroll: true
---

## Overview

This is a quality control report for the processing of MALDI-IHC-imaging datasets using the gutenTAG pipeline.

The following are reported on:

-   Mean-variance plots
-   Signal-noise analysis
-   Spatial QC

Load inputs

```{r Load params, echo=FALSE}
#param_load_dev_path <- params$param_load_dev_path
```

```{r Load gutenTAG, message=FALSE}
#if (!is.na(param_load_dev_path) && param_load_dev_path != "NA") {
#  message("Loading development version of gutenTAG")
#  devtools::load_all(param_load_dev_path)
#} else {
#  library(gutenTAG)
#}

library(gutenTAG)
library(ggplot2)
library(plotly)
library(ggrepel)
library(sessioninfo)
library(mclust)
library(tidyverse)

loadRData <- function(fileName){
  load(fileName, temp_env <- new.env())
  return(as.list(temp_env))
}

test_results <- loadRData("extdata/Example_processed.Rdata")
results <- test_results$results
metapeaks <- results$metapeaks
processed <- results$processed
panel <- results$panel
```

## Metapeak-level QC

### Metapeak Segmentation

It is very important to check the metapeak segmentation itself. We can visualise this by plotting out the counts histogram with the smooth counts and the metapeak limits overlaid on top.

```{r Dataframe for metapeak plotting}
# make dataframe for plotting test metapeaks
metapeak_df <- data.frame(matrix(0, ncol = 4, nrow = length(metapeaks$metapeaks$center)))
colnames(metapeak_df) <- c("center", "max", "lower_limit", "upper_limit")

metapeak_df$center <- metapeaks$metapeaks$center
metapeak_df$max <- metapeaks$metapeaks$max
metapeak_df$lower_limit <- metapeaks$metapeaks$limits[, 1]
metapeak_df$upper_limit <- metapeaks$metapeaks$limits[, 2]

max_y <- max(metapeaks$count_df$count, na.rm = TRUE)

# which metapeaks are in the targeted list
targeted_metapeaks <- processed$CorrespondenceMatrix$mz_location
targeted_metapeak_df <- metapeak_df[metapeak_df$max %in% targeted_metapeaks, ]

```

```{r Plot current metapeaks, out.width="100%", warning = FALSE}

# plot current metapeaks
# Create the ggplot object for counts (left axis) only
base_plot <- ggplot() +
  theme_minimal() +
  ggtitle("Metapeaks") +
  
  # Plot counts (left y-axis)
  geom_line(data = metapeaks$count_df, aes(x = mz, y = count, color = "Count"), alpha = 1, linewidth = 0.5) +
  # 8 is a really arbitrary number but it tends to put the counts in a more reasonable range for visualisation
  geom_line(data = metapeaks$count_smooth_df, aes(x = mz, y = count / 8, color = "SmoothCount"), alpha = 1, linewidth = 0.5) +
  #geom_line(data = metapeaks$count_smooth_df, aes(x = mz, y = count, color = "SmoothCount"), alpha = 1, linewidth = 0.5) +
  
  
  # Add vertical lines and ranges
  geom_vline(data = panel, aes(xintercept = FeatureMass, linetype = "Feature Mass", text = Name), color = 'purple', linewidth = 1) +
  geom_vline(data = metapeak_df, aes(xintercept = max, linetype = "Max Counts"), color = 'springgreen3', linewidth = 1) +
  #geom_vline(aes(xintercept = metapeaks_011$seed_mz, linetype = "Seeds"), color = 'red', linewidth = 0.5) +
  geom_vline(data = metapeak_df, aes(xintercept = metapeaks$seed_mz, linetype = "Seeds"), color = 'turquoise', linewidth = 1) +
  
  geom_rect(data = metapeak_df, aes(xmin = lower_limit, xmax = upper_limit, ymin = 0, ymax = max_y, fill = "Metapeak Range"), alpha = 0.2) +
  geom_rect(data = targeted_metapeak_df, aes(xmin = lower_limit, xmax = upper_limit, ymin = 0, ymax = max_y, fill = "Targeted Metapeak Range"), alpha = 0.2) +

  # Labels and legends
  ylab('Peak Counts') +
  scale_color_manual(name = "Line", values = c("Count" = "black", "SmoothCount" = "sienna")) +
  scale_linetype_manual(name = "Vertical Lines", values = c("Feature Mass" = "", "Max Counts" = "dashed", "Seeds" = "dashed")) +
  scale_fill_manual(name = "Range Areas", values = c("Metapeak Range" = 'grey', "Targeted Metapeak Range" = "red3"))



# Convert ggplot to plotly (without the spectra)
ggplotly_plot <- ggplotly(base_plot, dynamicTicks = TRUE)

# add mean and skyline spectra
ggplotly_plot_final <- ggplotly_plot |>
  add_trace(
    data = processed$SummarySpectra,
    x = ~mz,
    y = ~mean,
    type = 'scatter',
    mode = 'lines',
    name = 'MeanSpectrum',
    yaxis = 'y2',          # Tie this trace to the second y-axis
    line = list(color = '#cd0000', width = 2)
  ) |>
   add_trace(
    data = processed$SummarySpectra,
    x = ~mz,
    y = ~skyline,
    type = 'scatter',
    mode = 'lines',
    name = 'SkylineSpectrum',
    yaxis = 'y2',          # Tie this trace to the second y-axis
    line = list(color = "blue", width = 2)
  ) |>
  layout(
    yaxis = list(title = 'Peak Counts',
                 titlefont = list(size = 17, color = 'black'),
                 tickfont = list(size = 14, color = 'black')),          # Left y-axis (counts)
    yaxis2 = list(
      title = 'Spectral Intensity',               # Right y-axis (intensity)
      overlaying = 'y',                           # Overlay on left y-axis
      side = 'right',                             # Position on right side
      showgrid = FALSE,                           # No grid for right y-axis
      titlefont = list(size = 17, color = '#cd0000'),
      tickfont = list(size = 14, color = '#cd0000'),
      tickpadding = 0),
    margin = list(r = 1),                        # Increase right margin
    #legend = list(title = list(text = "Legend")),
    legend = list(
      x = 1.15,                # move legend further right if desired
      y = 1,
      title = list(text = "Legend"),
      showlegend = TRUE,
      font = list(size = 8)
    )                       
  )

# Render the plot
ggplotly_plot_final

```


## Feature-level QC

Current feature-level quality control metrics include:

 - Mean-variance relationship
 - Signal-to-Noise ratio (SNR)
 - Geary's C autocorrelation measure
 
These measures can be helpful in determining which features should be kept or removed during feature selection.

### Mean-Signal

As a simple first step, let's look at the mean signal strength from different markers.

```{r SNR barchart, eval=TRUE, warning=FALSE, message=FALSE}

# arrange by snr
mean_df <- dplyr::arrange(processed$CorrespondenceMatrix, mean)

df <- processed$IntensityDF

df_long <- df %>%
  pivot_longer(
    cols = everything(),
    names_to = "marker",
    values_to = "intensity"
  )

# optional: remove zeros if they represent off-tissue / no signal
#df_long <- df_long %>%
#  filter(intensity > 0)

# violin plot
ggplot(df_long, aes(x = reorder(marker, -intensity), y = intensity)) +
  geom_violin(
    fill = "grey80",
    color = "black",
    scale = "width",
    trim = TRUE
  ) +
  geom_boxplot(
    width = 0.15,
    outlier.shape = NA,        # don't double-plot outliers
    fill = "white",
    color = "black"
  ) +
  theme_minimal() +
  labs(
    x = "",
    y = "Mean intensity per pixel",
    title = "Mean Marker Intensities"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

### Mean-Variance relationship

The relationship between the mean and variance is of high importance when dealing with count data. Count data is often modelled using the Poisson distribution, which states that the mean and variance are equal. Real world count data is often over- or under-dispersed, meaninf that variance is greater or less than the mean, respectively. This can be easily visualised using a mean-variance plot. Here we show the relationship between the mean and variance of each metapeak. In red are annotated metapeaks (assigned to a target mass in the panel), while in grey are unassigned metapeaks.

Why is this important? In analysis of high dimensional biological datasets, it is common to take the highest variance genes or proteins and call these "interesting", because they are highly variable. However, since the mean and variance generally track together in count data, the highest intensity proteins will also be the highest variance ones. This can be corrected for by regressing the mean against the variance and extracting the residuals from this regression. These residuals tell us how the each marker deviates from the expected variance under poisson assumptions. Here, the x and y axes are on log scale.


```{r Mean-var plot data loading, message=FALSE, warning=FALSE}

# mean and sd of untarted metapeaks
untargeted_df <- data.frame(processed$Untargeted$UntargetedIntensity)

untargeted_mean <- colMeans(untargeted_df)
untargeted_sd = apply(untargeted_df, MARGIN = 2, FUN = sd)

# mean and sd of targeted dataframe
targeted_mean <- colMeans(processed$IntensityDF)
targeted_sd = apply(processed$IntensityDF, MARGIN = 2, FUN = sd)

targeted_stats <- data.frame(cbind(targeted_mean, targeted_sd))
targeted_stats <- cbind(targeted_stats, rownames(targeted_stats))
colnames(targeted_stats)[3] <- "Name"
untargeted_stats <- data.frame(cbind(untargeted_mean, untargeted_sd))

# linear model of mean vs variance and get coefficients
regression <- lm(log10(untargeted_sd) ~ log10(untargeted_mean))
coefs <- regression$coefficients
coefs <- data.frame(t(coefs))
colnames(coefs)[1] <- "Intercept"
colnames(coefs)[2] <- "Slope"

# plots
scatter <- ggplot() +
    theme_minimal() +
    ggtitle("Mean-Variance Plot") +
    geom_point(data = untargeted_stats, aes(x = untargeted_mean, y = untargeted_sd), color = 'black', fill = 'grey', shape = 21, size = 2) +
    geom_point(data = targeted_stats, aes(x = targeted_mean, y = targeted_sd, text = paste0("Marker: ", Name)), color = 'black', fill = 'red3', shape = 21, size = 3) +
    scale_x_log10() +
    scale_y_log10()

annotated <- ggplot() +
    theme_minimal() +
    ggtitle("Mean-Variance Plot: Annotated Peaks only") +
    geom_point(data = targeted_stats, aes(x = targeted_mean, y = targeted_sd, text = Name), color = 'black', fill = '#3B9AB5', shape = 21, size = 4) +
    scale_x_log10() +
    scale_y_log10()

```

::: {.panel-tabset group="language"}
## All Metapeaks

```{r , warning = FALSE, out.width="100%", echo = FALSE, message=FALSE}
ggplotly(scatter, dynamicTicks = TRUE)
```

## Annotated metapeaks

```{r , warning = FALSE, out.width="100%", echo = FALSE, message=FALSE}
ggplotly(annotated, dynamicTicks = TRUE)
```

:::

### Geary's C: Spatial QC

Geary's C score is a spatial autocorrelation score borrowed from Geostatistics. It is a metric to show on average how correlated spatially neighbouring pixels are to one another. A score of 0 can be interpreted as total spatial correlation, i.e. neighbouring pixels are similar to each other and are far away from dissimilar pixels. Scores close to 0 imply a high degree of structure in the single channel ion image. A score of 1 can be interpreted as spatial randomness, i.e there is no correlation between neighbouring pixels. Often markers with Geary's C scores close to one are unstructured, which can be as a result of poor staining, or some other underlying technical error. It can also be that unannotated metapeaks have a higher Geary's C score. The score can also extend above 1, implying negative spatial autocorrelation i.e. avoidance, where neighbouring pixels are dissimilar to each other.

In all Geary's C score is a handy statistic to show how spatially structured the ion images for each marker are. Spatially structured images are usually indicative of high quality staining.

```{r Gearys C barchart}

processed_geary <- computeGearysC(processed, update_correspondence = TRUE)

geary_sorted_df <- processed_geary$CorrespondenceMatrix[order(processed_geary$CorrespondenceMatrix$GearysC, decreasing = FALSE),]

marker_vector <- geary_sorted_df$marker[!is.na(geary_sorted_df$GearysC)]
geary_vector <- geary_sorted_df$GearysC[!is.na(geary_sorted_df$GearysC)]
geary_mean <- mean(geary_vector)

barplot(geary_vector,
        names.arg = marker_vector, cex.names = 0.4, las = 2,
        ylim = c(0,1),
        main = "Geary's C score", ylab = " Geary's C",
        col = "#b51837")
abline(h = geary_mean, col = "grey", lty = 2)

# Add a horizontal line for the mean Geary's C
abline(h = geary_mean, col = "grey", lty = 2)


```


### SNR

Signal-to-noise ratio can be a good indicator of how strong the observed signal is for a given channel. This can be a function of antibody staining and marker abundance, among other factors.

It may be worth considering removing features that perform poorly in SNR.

Function for computing SNR

```{r}
computeSNR <- function(x, package = "mclust", update_correspondence = FALSE, q = 1) {

  # Extract data & reorder dataframe by mz order to match correspondence
  df <- x$IntensityDF[, x$CorrespondenceMatrix$marker]

  # quantile clipping
  df <- apply(df, MARGIN = 2, FUN = .quantileClipping, q = q)

  # Apply SNR calculation based on the chosen method
  if (package == "mclust") {
    sample_snrs <- apply(df, MARGIN = 2, FUN = .snr_calc_mclust)
  } else if (package == "flexmix") {
    sample_snrs <- apply(df, MARGIN = 2, FUN = .snr_calc_flexmix)
  } else {
    stop("Package must be either 'mclust' or 'flexmix'")
  }

  # Update correspondence matrix if requested
  if (update_correspondence) {
    updated <- .updateCorrespondence(x, sample_snrs)
    return(updated)
  } else {
    return(sample_snrs)
  }
}

# Helper function for updating correspondence matrix
.updateCorrespondence <- function(x, sample_snrs) {
  snrs <- sapply(sample_snrs, function(y) y$SNR)
  x$CorrespondenceMatrix$snr <- snrs
  x$SNR <- sample_snrs
  return(x)
}

.quantileClipping <- function(x, q){
  quant <- quantile(x, q)
  x[x > quant] <- quant
  return(x)
}

# SNR calculation with mclust
.snr_calc_mclust <- function(channel) {
  # Remove zero values; if channel is all zeros, return an SNR of 0
  nonzero_channel <- channel[channel != 0]
  if (length(nonzero_channel) == 0) {
    message("The channel contains only zeros. SNR set to 0.")
    return(list("SNR" = 0, "clustering" = NULL, "fit" = NULL))
  }

  # Fit a GMM with mclust's Mclust() using two components
  fit <- Mclust(log2(nonzero_channel + 1), G = 2, modelNames = "E", verbose = FALSE)  # "E" for equal variance

  if (is.null(fit)) {
    stop("The model failed to converge. Try different data or check your input.")
  }

  clusters <- fit$classification
  cluster_means <- tapply(nonzero_channel, clusters, mean)

  # Identify noise and signal clusters based on the means
  noise_cluster <- which.min(cluster_means)
  signal_cluster <- which.max(cluster_means)

  # Reassign points in signal cluster below noise mean
  threshold <- min(nonzero_channel[clusters == signal_cluster])
  clusters <- ifelse(nonzero_channel < threshold, noise_cluster, clusters)

  # Recalculate means after reassignment
  noise_mean <- mean(nonzero_channel[clusters == noise_cluster])
  signal_mean <- mean(nonzero_channel[clusters == signal_cluster])
  snr <- signal_mean / noise_mean

  return(list("SNR" = snr, "clustering" = clusters, "fit" = fit))
}

# SNR calculation with flexmix
.snr_calc_flexmix <- function(channel) {
  nonzero_channel <- channel[channel != 0]
  if (length(nonzero_channel) == 0) {
    message("The channel contains only zeros. SNR set to 0.")
    return(list("SNR" = 0, "clustering" = NULL, "fit" = NULL))
  }

  data <- data.frame(nonzero_channel)
  fit <- tryCatch({
    flexmix(log2(nonzero_channel + 1) ~ 1, data = data, k = 2,
            model = FLXMRglm(family = "gaussian"),
            control = list(iter.max = 500, tol = 1e-6))
  }, error = function(e) {
    message("Error in flexmix fitting: ", e)
    return(NULL)
  })

  if (is.null(fit)) {
    stop("The model failed to converge.")
  }

  clusters <- clusters(fit)
  cluster_means <- tapply(nonzero_channel, clusters, mean)

  # Identify noise and signal clusters based on the means
  noise_cluster <- which.min(cluster_means)
  signal_cluster <- which.max(cluster_means)

  # Reassign points in signal cluster below noise mean
  threshold <- min(nonzero_channel[clusters == signal_cluster])
  clusters <- ifelse(nonzero_channel < threshold, noise_cluster, clusters)

  # Recalculate means after reassignment
  noise_mean <- mean(nonzero_channel[clusters == noise_cluster])
  signal_mean <- mean(nonzero_channel[clusters == signal_cluster])
  snr <- signal_mean / noise_mean

  return(list("SNR" = snr, "clustering" = clusters, "fit" = fit))
}


```


```{r SNR barchart, eval=TRUE, warning=FALSE, message=FALSE}

processed_snr <- computeSNR(processed_geary, package = "mclust", update_correspondence = TRUE, q = 0.995)
filtered_processed_snr <- processed_snr$CorrespondenceMatrix[!is.na(processed_snr$CorrespondenceMatrix$mz_location), ]

# arrange by snr
snr_barplot_df <- dplyr::arrange(filtered_processed_snr, snr)

# set value for recommended snr cutoff threshold, currently 10th percentile
threshold <- quantile(snr_barplot_df$snr, 0.1)

# plot
snr_plot <- ggplot(snr_barplot_df, aes(x = reorder(marker, -snr), y = snr, fill = snr <= threshold)) +
  geom_bar(stat = "identity") +
  
  # Custom fill colors based on the SNR values
  scale_fill_manual(values = c("#18B596", "#b51837")) +
  
  # Add horizontal lines
  geom_hline(yintercept = threshold, linetype = "dashed", color = "black") +
  
  # Add text labels near the horizontal lines
  annotate("text", x = 8.3, y = threshold + 0.03, label = "Filter: 10th percentile", color = "black", hjust = 0, vjust = -0.5, size = 3) +

  
  # Customizations
  labs(title = "SNR-based Filtering", x = NULL, y = "SNR") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 5))

# Display the plot
print(snr_plot)

```

### SNR Distribution

It is important to see the distribution of the signal and noise clusters on the intensity histogram for each channel to fully trust the SNR calculation. Note that due to low number of pixels in this example image, the histograms are quite sporadic.

::: {.panel-tabset group="snr"}

```{r snr plotting, out.width="100%", results='asis'}

for (i in 1:length(processed_snr$SNR)) {
  
  # Skip channels with a mean of zero
  if (mean(processed_snr$IntensityDF[, i]) == 0) {
    next
  }

  # Extract necessary data for the current channel
  coords <- processed$SpatialCoords[processed$IntensityDF[, i] != 0, ]
  cur_name <- colnames(processed_snr$IntensityDF)[i]
  channel <- processed_snr$IntensityDF[, i]
  nonzero_channel <- channel[channel != 0]
  
  # Cap the intensities at a certain quantile (if needed)
  quant <- quantile(nonzero_channel, 0.999) # Adjust quantile as needed
  nonzero_channel[nonzero_channel > quant] <- quant

  # Retrieve current clustering information
  cur_cluster <- processed_snr$SNR[[i]]$clustering
  cur_snr <- processed_snr$SNR[[i]]$SNR

  # Identify noise and signal based on cluster means
  cluster_means <- tapply(nonzero_channel, cur_cluster, mean)
  noise_mean <- min(cluster_means)
  signal_mean <- max(cluster_means)
  cluster_color <- ifelse(cur_cluster == which.min(cluster_means), 'Noise', 'Signal')

  # Data for histogram
  hist_df <- data.frame(
    count = nonzero_channel,
    cluster_color = cluster_color
  )

  # Generate histogram plot for the current channel
  histogram_plot <- ggplot(hist_df, aes(x = count, fill = cluster_color)) +
    geom_histogram(bins = 100, color = "black", alpha = 0.5, position = "identity") +
    scale_fill_manual(values = c('Noise' = 'grey', 'Signal' = 'red')) + 
    theme_minimal() +
    labs(
      title = paste("Histogram of Signal and Noise: ", cur_name),
      subtitle = paste("Signal-to-Noise Ratio (SNR): ", round(cur_snr, 2)),
      x = "Intensity", y = "Count", fill = "Cluster"
    ) +
    #scale_x_log10()
  
  # Dynamically create the tab headers and content
  cat("##", cur_name, "\n\n")
  print(histogram_plot)
  cat("\n\n")
}


```

:::

It can also be important to see the spatial distribution of the signal and noise clusters to validate that this corresponds to the ion image for the given channel. Also note that due to low pixel numbers, these images appear non-compact with this method of illustration. 

::: {.panel-tabset}

```{r spatial snr, out.width = "100%", results='asis'}
for (i in 1:length(processed_snr$SNR)) {
  
  # Skip channels with a mean of zero
  if (mean(processed_snr$IntensityDF[, i]) == 0) {
    next
  }

  # Extract necessary data for the current channel
  coords <- processed$SpatialCoords[processed$IntensityDF[, i] != 0, ]
  cur_name <- colnames(processed_snr$IntensityDF)[i]
  channel <- processed_snr$IntensityDF[, i]
  nonzero_channel <- channel[channel != 0]
  
  # Retrieve current clustering information
  cur_cluster <- processed_snr$SNR[[i]]$clustering
  cur_snr <- processed_snr$SNR[[i]]$SNR

  # Identify noise and signal based on cluster means
  cluster_means <- tapply(nonzero_channel, cur_cluster, mean)
  noise_mean <- min(cluster_means)
  signal_mean <- max(cluster_means)
  cluster_color <- ifelse(cur_cluster == which.min(cluster_means), 'Noise', 'Signal')

  # Prepare the dataframe for spatial plotting
  plot_df <- cbind(coords, cluster_color)
  
  # Generate spatial plot for the current channel
  spatial_plot <- ggplot(plot_df, aes(x = x, y = -y, color = factor(cluster_color))) +
    geom_point(size = 7) +
    scale_color_manual(values = c('Noise' = 'grey', 'Signal' = 'red')) + 
    theme_void() +
    labs(
      title = paste("Spatial Distribution: ", cur_name),
      x = "X", y = "Y", color = "Cluster"
    )
  
  # Dynamically create the tab headers and content
  cat("##", cur_name, "\n\n")
  print(spatial_plot)
  cat("\n\n")
}

```

### Mean Signal vs SNR plot

Let's have a look at the relationship between mean signal strength and SNR.

```{r mean_vs_snr}

# mean and SNR of targeted dataframe
targeted_mean <- colMeans(processed$IntensityDF)
targeted_snr <- processed_snr$CorrespondenceMatrix$snr

targeted_mean_snr <- data.frame(cbind(targeted_mean, targeted_snr))
targeted_mean_snr <- cbind(targeted_mean_snr, rownames(targeted_mean_snr))
colnames(targeted_mean_snr)[3] <- "Name"

# linear model of mean vs variance and get coefficients
regression <- lm(log10(targeted_snr) ~ log10(targeted_mean))
coefs <- regression$coefficients
coefs <- data.frame(t(coefs))
colnames(coefs)[1] <- "Intercept"
colnames(coefs)[2] <- "Slope"

# plots
scatter <- ggplot() +
    theme_minimal() +
    ggtitle("Mean-Variance Plot") +
    geom_point(data = targeted_mean_snr, aes(x = targeted_mean, y = targeted_snr), color = 'black', fill = 'grey', shape = 21, size = 2) +
    geom_point(data = targeted_stats, aes(x = targeted_mean, y = targeted_snr, text = paste0("Marker: ", Name)), color = 'black', fill = 'red3', shape = 21, size = 3) +
    scale_x_log10() +
    scale_y_log10()

annotated <- ggplot() +
    theme_minimal() +
    ggtitle("Mean-Variance Plot: Annotated Peaks only") +
    geom_point(data = targeted_mean_snr, aes(x = targeted_mean, y = targeted_snr, text = Name), color = 'black', fill = '#3B9AB5', shape = 21, size = 4)# +
    #scale_x_log10() +
    #scale_y_log10()

print(annotated)

```


:::

### Geary-SNR plot: Feature Selection

Channels with a high Geary's C score and low SNR score can be confidently determined to be weak channels. It can be important to look at both of these metrics to avoid abberant removal of markers. Consider the case of a marker with a strong SNR, but with very uniform staining due to the ubiquity of the marker in the sample. Using only Geary's C for feature selection may result in identifying this marker as performing poorly.

```{r trashcan plot, eval=TRUE, message=FALSE, warning=FALSE}

# fixed thresholds
geary_threshold <- 0.7
snr_threshold <- 3

# conditions for fill
geary_condition <- filtered_processed_snr$GearysC > geary_threshold
snr_condition <- filtered_processed_snr$snr < snr_threshold
filter <- geary_condition & snr_condition
filtered_processed_snr$fill_color <- ifelse(filter, "#b51837", "#18B596")

# Define the trashcan condition: SNR ≤ 2 and Geary's C ≥ 0.8
trashcan_zone <- filtered_processed_snr |>
  dplyr::filter(snr <= snr_threshold & GearysC >= geary_threshold)

all_markers <- filtered_processed_snr

# Create the trashcan plot with labels for markers in the trashcan zone
trashcan_plot <- ggplot() +
  theme_minimal() +
  ggtitle("Mean-SNR Plot: Trashcan plot") +
  # Shaded red rectangle (alpha = 0.2) above the vertical line and below the horizontal line
  geom_rect(aes(xmin = geary_threshold, xmax = 1, ymin = 0, ymax = snr_threshold), 
            fill = "red3", alpha = 0.5) +
  # Scatter plot for all markers
  geom_point(data = filtered_processed_snr, aes(x = GearysC, y = snr, fill = fill_color, text = marker), 
             color = 'black', shape = 21, size = 4) +
  # Horizontal and vertical lines
  geom_hline(yintercept = snr_threshold, linetype = "solid", color = "black") +
  geom_vline(xintercept = geary_threshold, linetype = "solid", color = "black") +
  # Labels for points in the trashcan zone (markers with SNR ≤ 2 and Geary's C ≥ 0.8)
  geom_label_repel(data = all_markers, aes(x = GearysC, y = snr, label = marker),
                   color = 'black',   # Border color
                   fill = 'white',    # Background color (inside the label)
                   box.padding = 0.2, 
                   point.padding = 0.2, 
                   max.overlaps = Inf, 
                   segment.color = 'black', # Line connecting the label to the point
                   size = 2, fontface = 'bold', label.size = 0.2) +  # label.size controls the border thickness
  # Use the fill color as defined in the data without further scaling
  #scale_x_log10() +
  scale_y_log10() +
  scale_fill_identity()


```

::: {.panel-tabset group="plottype"}
## Simple

```{r , warning = FALSE, out.width="100%", echo = FALSE, message=FALSE}
print(trashcan_plot)
```

## Interactive

```{r , warning = FALSE, out.width="100%", echo = FALSE, message=FALSE}
ggplotly(trashcan_plot, dynamicTicks = FALSE)
```

:::


Here is the list of markers that perform poorly according to this plot, and are worth considering to remove from the feature list.

```{r print trashcan markers, eval=TRUE}

dput(sort(trashcan_zone$marker))

```

## Pixel-level QC

### TIC histogram thresholding

Thresholding on the total ion current of the sample can be potentially useful in terms of determining which pixels are off-tissue regions. Off-tissue pixels often have a lower TIC than tissue pixels.

```{r TIC histogram thresholding}

# Compute histogram
tic_targeted <- rowSums(processed$IntensityDF)
tic_targeted[tic_targeted == 0] <- NA
tic_targeted <- na.omit(tic_targeted)

threshold <- gutenTAG:::.otsu_thresholding(log10(tic_targeted))

# Compute the histogram data without plotting it
normalised_histogram <- hist(log10(1 + tic_targeted), 100, plot = FALSE)

# Plot the histogram with custom colors
bar_colors <- ifelse(normalised_histogram$mids <= threshold, "grey", "red3")

# Plot the histogram
plot(normalised_histogram, col = bar_colors, border = "black",
     main = "Normalised TIC histogram", cex.main = 2,
     xlab = "Total Ion Count (TIC)",
     xaxs = 'i', yaxs = 'i',
     cex.lab = 1.3)

# Add the threshold line
abline(v = threshold, lwd = 4, col = "grey", lty = 2)

```


```{r , eval=FALSE, echo=FALSE}

# Identify noise and signal based on cluster means
cluster_color <- ifelse(log10(1 + tic_targeted) <= threshold, "Noise", "Signal")

# Prepare the dataframe for spatial plotting
plot_df <- cbind(coords, cluster_color)

# Generate spatial plot for the current channel
spatial_plot <- ggplot(plot_df, aes(x = x, y = -y, color = factor(cluster_color))) +
  geom_point(size = 7) +
  scale_color_manual(values = c('Noise' = 'grey', 'Signal' = 'red')) + 
  theme_void() +
  labs(
    title = paste("Spatial Distribution: ", cur_name),
    x = "X", y = "Y", color = "Cluster"
  )

print(spatial_plot)

```


:::{.callout-tip collapse="true"}
## Expand for Session Info
```{r, echo = FALSE}
library(sessioninfo)
# save the session info as an object
pkg_sesh <- session_info(pkgs = "attached")

# get the quarto version
quarto_version <- system("quarto --version", intern = TRUE)

# inject the quarto info
pkg_sesh$platform$quarto <- paste(
  system("quarto --version", intern = TRUE), 
  "@", 
  quarto::quarto_path()
  )

# print it out
pkg_sesh
```

:::
