---
title: "Quality Control Report"
params:
    param_load_dev_path: NA
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    toc-expand: 1
    toc-title: Contents
    toc-location: right
    code-fold: true
    code-tools: true
    code-summary: "Show the code"
    css: styles.css
    smooth-scroll: true
---

## Overview

This is a quality control report for the processing of MALDI-IHC-imaging datasets using the gutenTAG pipeline.

The following are reported on:

-   Mean-variance plots
-   Mass shift
-   Signal-noise analysis
-   Spatial QC

Load inputs

```{r Load gutenTAG, message=FALSE, warning=FALSE}

library(gutenTAG)
library(plotly)
library(ggrepel)
library(sessioninfo)

loadRData <- function(fileName){
  load(fileName, temp_env <- new.env())
  return(as.list(temp_env))
}

test_results <- loadRData("extdata/Example_processed.Rdata")
results <- test_results$results
metapeaks <- results$metapeaks
processed <- results$processed
panel <- results$panel
```

## Metapeak-level QC

### Metapeak Segmentation

It is very important to check the metapeak segmentation itself. We can visualise this by plotting out the counts histogram with the smooth counts and the metapeak limits overlaid on top.

```{r Plot current metapeaks, out.width="100%", warning=FALSE}

plotMetapeaks(x = processed, metapeaks = metapeaks, panel = panel,
              interactive = TRUE)

```

The dashed horizontal line marks the detection threshold (the minimum pixel count required for a metapeak to be retained).

### Mass Shift

Each targeted metapeak is picked at the m/z location with the highest peak count, which may differ slightly from the expected mass defined in the panel. This deviation — the mass shift — is a useful QC metric. a tight distribution centred near zero indicates a small degree of mass shift, while a wide spread or a systematic offset can point to calibration issues. If you observe most markers are above/below the zero line, while some are in the opposite direction, these could be falsely assigned and are worth looking into further. We compute the shift as the picked m/z minus the expected m/z.

The distribution view shows the overall spread of the mass shift across all targeted peaks; the spectrum view plots the same shift against the expected m/z, so any systematic drift along the mass range becomes visible.

::: {.panel-tabset group="massshift"}
## Distribution

```{r mass_shift_distribution, out.width="100%", warning=FALSE, message=FALSE}
plotMassShift(x = processed, type = "distribution", 
              mean_point = list(color = "black", fill = "black", alpha = 1, shape = 22, size = 8),
              line = list(linetype = "dashed", color = "black"), 
              points = list(color = "black", fill = "orange", alpha = 0.9, size = 5)
              )
```

## Across m/z

```{r mass_shift_spectrum, out.width="100%", warning=FALSE, message=FALSE}
plotMassShift(x = processed, type = "spectrum",
              mean_point = list(color = "black", fill = "black", alpha = 1, shape = 22, size = 8),
              line = list(linetype = "dashed", color = "black"), 
              points = list(color = "black", fill = "orange", alpha = 0.9, size = 5))
```
:::

## Feature-level QC

Current feature-level quality control metrics include:

-   Mean-variance relationship
-   Signal-to-Noise ratio (SNR)
-   Geary's C autocorrelation measure

These measures can be helpful in determining which features should be kept or removed during feature selection.

### Mean Marker Signal

As a simple first step, let's look at the mean signal strength from different markers.

```{r mean_intensity_violin, eval=TRUE, warning=FALSE, message=FALSE}

plotIntensityDistribution(x = processed, boxplot = list(fill = "orange", alpha = 0.6),
                          violin = list(fill = "orange", alpha = 0.6))

```

### Mean-Variance relationship

The relationship between the mean and variance is of high importance when dealing with count data. Count data is often modelled using the Poisson distribution, which states that the mean and variance are equal. Real world count data is often over- or under-dispersed, meaninf that variance is greater or less than the mean, respectively. This can be easily visualised using a mean-variance plot. Here we show the relationship between the mean and variance of each metapeak. In red are annotated metapeaks (assigned to a target mass in the panel), while in grey are unassigned metapeaks.

Why is this important? In analysis of high dimensional biological datasets, it is common to take the highest variance genes or proteins and call these "interesting", because they are highly variable. However, since the mean and variance generally track together in count data, the highest intensity proteins will also be the highest variance ones. This can be corrected for by regressing the mean against the variance and extracting the residuals from this regression. These residuals tell us how the each marker deviates from the expected variance under poisson assumptions. Here, the x and y axes are on log scale.

::: {.panel-tabset group="language"}
## All Metapeaks

```{r mean_var_all, warning=FALSE, out.width="100%", message=FALSE}
plotMeanVariance(x = processed, interactive = FALSE, annotated_only = FALSE, 
                 points = list(fill = "orange", color = "black", size = 4), 
                 unannotated_points = list(fill = "grey"),
                 regression = list(color = "orange", linetype = "dashed", method = "loess", se = TRUE), 
                 show_stats = TRUE, 
                 stats = list(color = "black")
                 )
```

## Annotated metapeaks

```{r mean_var_annotated, warning=FALSE, out.width="100%", message=FALSE}
plotMeanVariance(x = processed, annotated_only = TRUE, interactive = TRUE, points = list(fill = "orange", color = "black", size = 4), 
                 unannotated_points = list(fill = "grey"),
                 regression = list(color = "orange", linetype = "dashed", method = "lm", se = TRUE))
```

## Residuals

```{r mean_var_residuals, warning=FALSE, out.width="100%", message=FALSE}
plotMeanVarianceResiduals(x = processed, interactive = TRUE, 
                          line = list(color = "red", linewidth = 0.5, linetype = "dashed"),
                          points = list(fill = "orange"))
```

## Standardised Residuals

```{r mean_var_std_residuals, warning=FALSE, out.width="100%", message=FALSE}
plotMeanVarianceResiduals(x = processed, standardised = TRUE, interactive = TRUE,
                          line = list(color = "red", linewidth = 0.5, linetype = "dashed"),
                          points = list(fill = "orange"))
```
:::

### Geary's C: Spatial QC

Geary's C score is a spatial autocorrelation score borrowed from Geostatistics. It is a metric to show on average how correlated spatially neighbouring pixels are to one another. A score of 0 can be interpreted as total spatial correlation, i.e. neighbouring pixels are similar to each other and are far away from dissimilar pixels. Scores close to 0 imply a high degree of structure in the single channel ion image. A score of 1 can be interpreted as spatial randomness, i.e there is no correlation between neighbouring pixels. Often markers with Geary's C scores close to one are unstructured, which can be as a result of poor staining, or some other underlying technical error. It can also be that unannotated metapeaks have a higher Geary's C score. The score can also extend above 1, implying negative spatial autocorrelation i.e. avoidance, where neighbouring pixels are dissimilar to each other.

In all Geary's C score is a handy statistic to show how spatially structured the ion images for each marker are. Spatially structured images are usually indicative of high quality staining.

```{r gearys_c_barchart}

processed_geary <- computeGearysC(processed, update_correspondence = TRUE)

plotGearysC(x = processed_geary)
plotGearysC(x = processed_geary, bar = list(fill = "orange", color = "black"), line = list(color = "black", linewidth = 1))

```

### SNR

Signal-to-noise ratio can be a good indicator of how strong the observed signal is for a given channel. This can be a function of antibody staining and marker abundance, among other factors.

It may be worth considering removing features that perform poorly in SNR.

```{r snr_barchart, eval=TRUE, warning=FALSE, message=FALSE}

processed_snr <- computeSNR(processed_geary, package = "mclust",
                             update_correspondence = TRUE, q = 0.995)

plotSNR(x = processed_snr)
```

### SNR Distribution

It is important to see the distribution of the signal and noise clusters on the intensity histogram for each channel to fully trust the SNR calculation. Note that due to low number of pixels in this example image, the histograms are quite sporadic.

::: {.panel-tabset group="snr"}
```{r snr_histograms, out.width="100%", results='asis'}

for (i in seq_along(processed_snr$SNR)) {

  if (mean(processed_snr$IntensityDF[, i]) == 0) next

  cur_name <- colnames(processed_snr$IntensityDF)[i]
  cat("##", cur_name, "\n\n")
  print(plotSNRHistogram(x = processed_snr, channel = i))
  cat("\n\n")
}

```
:::

It can also be important to see the spatial distribution of the signal and noise clusters to validate that this corresponds to the ion image for the given channel. Also note that due to low pixel numbers, these images appear non-compact with this method of illustration.

::: panel-tabset
```{r spatial_snr, out.width="100%", results='asis'}

for (i in seq_along(processed_snr$SNR)) {

  if (mean(processed_snr$IntensityDF[, i]) == 0) next

  cur_name <- colnames(processed_snr$IntensityDF)[i]
  cat("##", cur_name, "\n\n")
  print(plotSpatialSNR(x = processed_snr, channel = i))
  cat("\n\n")
}

```

### Mean Signal vs SNR plot

Let's have a look at the relationship between mean signal strength and SNR.

```{r mean_vs_snr}

plotMeanVsSNR(x = processed_snr, points = list(color = "black", fill = "orange", alpha = 0.7), 
              regression = list(color = "black", linetype  = "dashed", se = TRUE))

```
:::

### QC overview plot: Feature Selection

Channels with a high Geary's C score and low SNR score can be confidently determined to be weak channels. It can be important to look at both of these metrics to avoid abberant removal of markers. Consider the case of a marker with a strong SNR, but with very uniform staining due to the ubiquity of the marker in the sample. Using only Geary's C for feature selection may result in identifying this marker as performing poorly.

::: {.panel-tabset group="plottype"}
## Simple

```{r trashcan_simple, warning=FALSE, out.width="100%", message=FALSE}
plotQCOverview(x = processed_snr, palette = c("pass" = "#18B596", "Noise" = "#b51837"))
```

## Interactive

```{r trashcan_interactive, warning=FALSE, out.width="100%", message=FALSE}
plotQCOverview(x = processed_snr, interactive = TRUE)
```
:::

Here is the list of markers that perform poorly according to this plot, and are worth considering to remove from the feature list.

```{r print trashcan markers, eval=TRUE}

cm <- processed_snr$CorrespondenceMatrix
cm <- cm[!is.na(cm$mz_location), ]
trashcan_zone <- cm[cm$snr <= 3 & cm$GearysC >= 0.7, ]
dput(sort(trashcan_zone$marker))

```

## Pixel-level QC

### TIC histogram thresholding

Thresholding on the total ion current of the sample can be potentially useful in terms of determining which pixels are off-tissue regions. Off-tissue pixels often have a lower TIC than tissue pixels.

```{r tic_histogram_thresholding}

plotTICHistogram(x = processed, 
                 histogram = list(bins = 20), vline = list(linetype = "dashed", linewidth = 2, color = "red2"),
                 palette = c(Signal = "orange", Noise = "grey80"))

```

```{r tic_spatial, eval=TRUE}

plotTICSpatial(x = processed, palette = c(Signal = "orange", Noise = "grey80"))

```

::: {.callout-tip collapse="true"}
## Expand for Session Info

```{r, echo=FALSE}
library(sessioninfo)
# save the session info as an object
pkg_sesh <- session_info(pkgs = "attached")

# get the quarto version
quarto_version <- system("quarto --version", intern = TRUE)

# inject the quarto info
pkg_sesh$platform$quarto <- paste(
  system("quarto --version", intern = TRUE),
  "@",
  quarto::quarto_path()
  )

# print it out
pkg_sesh
```
:::
