---
title: "Quality Control"
page-layout: full
params:
    param_load_dev_path: NA
---

## Overview

This is a quality control report for the processing of MALDI-IHC-imaging datasets using the gutenTAG pipeline.

The following are reported on:

-   Mean-variance plots
-   Signal-noise analysis
-   Spatial QC

### Load inputs

```{r Load params, echo=FALSE}
param_load_dev_path <- params$param_load_dev_path
```

```{r Load gutenTAG, message=FALSE}
if (!is.na(param_load_dev_path) && param_load_dev_path != "NA") {
  message("Loading development version of gutenTAG")
  devtools::load_all(param_load_dev_path)
} else {
  library(gutenTAG)
}

library(ggplot2)
library(plotly)

loadRData <- function(fileName){
  load(fileName, temp_env <- new.env())
  return(as.list(temp_env))
}

results <- loadRData("metapeak_result.RData")
metapeaks <- results$metapeaks
processed <- results$processed
panel <- results$panel
```

## Metapeak-level QC

### Metapeak Segmentation

It is very important to check the metapeak segmentation itself. We can visualise this by plotting out the counts histogram with the smooth counts and the metapeak limits overlaid on top.

```{r Dataframe for metapeak plotting}
# make dataframe for plotting test metapeaks
metapeak_df <- data.frame(matrix(0, ncol = 4, nrow = length(metapeaks$metapeaks$center)))
colnames(metapeak_df) <- c("center", "max", "lower_limit", "upper_limit")

metapeak_df$center <- metapeaks$metapeaks$center
metapeak_df$max <- metapeaks$metapeaks$max
metapeak_df$lower_limit <- metapeaks$metapeaks$limits[, 1]
metapeak_df$upper_limit <- metapeaks$metapeaks$limits[, 2]

max_y <- max(metapeaks$count_df$count, na.rm = TRUE)

# which metapeaks are in the targeted list
targeted_metapeaks <- processed$CorrespondenceMatrix$mz_location
targeted_metapeak_df <- metapeak_df[metapeak_df$max %in% targeted_metapeaks, ]

```

```{r Plot current metapeaks, out.width="100%", warning = FALSE}
# plot current metapeaks
# Create the ggplot object for counts (left axis) only
base_plot <- ggplot() +
  theme_minimal() +
  ggtitle("Metapeaks") +
  
  # Plot counts (left y-axis)
  geom_line(data = metapeaks$count_df, aes(x = mz, y = count, color = "Count"), alpha = 1, linewidth = 0.5) +
  
  # Add vertical lines and ranges
  geom_vline(data = panel, aes(xintercept = FeatureMass, linetype = "Feature Mass", text = Name), color = 'purple', linewidth = 0.7) +
  geom_vline(data = metapeak_df, aes(xintercept = max, linetype = "Max Counts"), color = 'springgreen3', linewidth = 0.5) +
  
  geom_rect(data = metapeak_df, aes(xmin = lower_limit, xmax = upper_limit, ymin = 0, ymax = max_y, fill = "Metapeak Range"), alpha = 0.2) +
  geom_rect(data = targeted_metapeak_df, aes(xmin = lower_limit, xmax = upper_limit, ymin = 0, ymax = max_y, fill = "Targeted Metapeak Range"), alpha = 0.2) +

  # Labels and legends
  ylab('Peak Counts') +
  scale_color_manual(name = "Line", values = c("Count" = "black")) +
  scale_linetype_manual(name = "Vertical Lines", values = c("Feature Mass" = "", "Max Counts" = "dashed")) +
  scale_fill_manual(name = "Range Areas", values = c("Metapeak Range" = 'grey', "Targeted Metapeak Range" = "red3"))


# Convert ggplot to plotly (without the spectra)
ggplotly_plot <- ggplotly(base_plot, dynamicTicks = TRUE)

# Add the spectra intensity as a separate trace (for right y-axis)
ggplotly_plot <- ggplotly_plot |>
  add_trace(
    data = processed$SummarySpectra,
    x = ~mz,
    y = ~mean,
    type = 'scatter',
    mode = 'lines',
    name = 'MeanSpectrum',
    yaxis = 'y2',          # Tie this trace to the second y-axis
    line = list(color = '#cd0000', width = 1.5)
  ) |>
  add_trace(
    data = processed$SummarySpectra,
    x = ~mz,
    y = ~skyline,
    type = 'scatter',
    mode = 'lines',
    name = 'SkylineSpectrum',
    yaxis = 'y2',          # Tie this trace to the second y-axis
    line = list(color = "blue", width = 1.5)
  ) |>
  
   #Customize layout for dual y-axes
  layout(
    yaxis = list(title = 'Peak Counts'),          # Left y-axis (counts)
    yaxis2 = list(
      title = 'Spectrum Intensity',               # Right y-axis (intensity)
      overlaying = 'y',                           # Overlay on left y-axis
      side = 'right',                             # Position on right side
      showgrid = FALSE                            # No grid for right y-axis
    ),
    legend = list(title = list(text = "Legend")),
    showlegend = TRUE
  )

# Render the plot
ggplotly_plot
```

## Feature-level QC

Current feature-level quality control metrics include:
 - Mean-variance relationship
 - Signal-to-Noise ratio (SNR)
 - Geary's C autocorrelation measure
 
These measures can be helpful in determining which features should be kept or removed during feature selection.

### Mean-Variance

The relationship between the mean and variance is of high importance when dealing with count data. Count data can be modeled simply using the Poisson distribution, which clearly states that the mean and variance are equal. Real world count data is often over- or under-dispersed. This means that variance is greater or less than the mean, respectively. This can be easily visualised using a mean-variance plot. Here we show the relationship between the mean and variance of each metapeak. In red are annotated metapeaks (assigned to a target mass in the panel), while in grey are unassigned metapeaks.

Why is this important? In analysis of high dimensional biological datasets, it is common to take the highest variance genes or proteins and call these "interesting", because they are highly variable. However, since the mean and variance generally track together in count data, the highest intensity proteins will also be the highest variance ones. This is the confounding effect of the mean. This can be corrected for by regressing the mean against the variance and extracting the residuals from this regression. These residuals tell us how the each marker deviates from the expected variance under poisson assumptions. Here, the x and y axes are on log scale.

#### Extract relevant data

```{r Mean-var plot data loading}

# mean and sd of untarted metapeaks
untargeted_df <- data.frame(processed$Untargeted$UntargetedIntensity)

untargeted_mean <- colMeans(untargeted_df)
untargeted_sd = apply(untargeted_df, MARGIN = 2, FUN = sd)

# mean and sd of targeted dataframe
targeted_mean <- colMeans(processed$IntensityDF)
targeted_sd = apply(processed$IntensityDF, MARGIN = 2, FUN = sd)

```

#### Mean-variance plot for annotated and unannotated metapeaks

##### Annotated and unannotated metapeaks

```{r ggplot mean var plot}

targeted_stats <- data.frame(cbind(targeted_mean, targeted_sd))
targeted_stats <- cbind(targeted_stats, rownames(targeted_stats))
colnames(targeted_stats)[3] <- "Name"
untargeted_stats <- data.frame(cbind(untargeted_mean, untargeted_sd))

# linear model of mean vs variance and get coefficients
regression <- lm(log10(untargeted_sd) ~ log10(untargeted_mean))
coefs <- regression$coefficients
coefs <- data.frame(t(coefs))
colnames(coefs)[1] <- "Intercept"
colnames(coefs)[2] <- "Slope"
```

```{r plot, warning = FALSE}
scatter <- ggplot() +
    theme_minimal() +
    ggtitle("Mean-Variance Plot") +
    geom_point(data = untargeted_stats, aes(x = untargeted_mean, y = untargeted_sd), color = 'black', fill = 'grey', shape = 21, size = 2) +
    geom_point(data = targeted_stats, aes(x = targeted_mean, y = targeted_sd, text = paste0("Marker: ", Name)), color = 'black', fill = 'red3', shape = 21, size = 3) +
    scale_x_log10() +
    scale_y_log10()

annotated <- ggplot() +
    theme_minimal() +
    ggtitle("Mean-Variance Plot: Annotated Peaks only") +
    geom_point(data = targeted_stats, aes(x = targeted_mean, y = targeted_sd, text = Name), color = 'black', fill = '#3B9AB5', shape = 21, size = 4) +
    scale_x_log10() +
    scale_y_log10()
```

```{r , warning = FALSE, out.width="100%", echo = FALSE}
ggplotly(scatter, dynamicTicks = TRUE)
ggplotly(annotated, dynamicTicks = TRUE)
```

### Geary's C: Spatial QC

Geary's C score is a spatial autocorrelation score borrowed from Geostatistics. It is a metric to show on average how correlated spatially neighbouring pixels are to one another. A score of 0 can be interpreted as total spatial correlation, i.e. neighbouring pixels are similar to each other and are far away from dissimilar pixels. Scores close to 0 imply a high degree of structure in the single channel ion image. A score of 1 can be interpreted as spatial randomness, i.e there is no correlation between neighbouring pixels. Often markers with Geary's C scores close to one are unstructured, which can be as a result of poor staining, or some other underlying technical error. It can also be that unannotated metapeaks have a higher Geary's C score. The score can also extend above 1, implying negative spatial autocorrelation i.e. avoidance, where neighbouring pixels are dissimilar to each other.

In all Geary's C score is a handy statistic to show how spatially structured the ion images for each marker are. Spatially structured images are usually indicative of high quality staining.

```{r Gearys C barchart}

processed_geary <- computeGearysC(processed, update_correspondence = TRUE)

geary_sorted_df <- processed_geary$CorrespondenceMatrix[order(processed_geary$CorrespondenceMatrix$GearysC, decreasing = FALSE),]
marker_vector <- geary_sorted_df$marker[!is.na(geary_sorted_df$GearysC)]
geary_vector <- geary_sorted_df$GearysC[!is.na(geary_sorted_df$GearysC)]
geary_mean <- mean(geary_vector)

#barplot(geary_vector,
#        names.arg = marker_vector, cex.names = 0.4, las = 2,
#        ylim = c(0,1),
#        main = "Geary's C score", ylab = " Geary's C",
#        col = "#b51837")
#abline(h = geary_mean, col = "grey", lty = 2)

# Set the default color for all bars as grey
bar_colors <- ifelse(marker_vector %in% c("ATP5a", "S100A4"), "#b51837", "grey")

# Create the barplot with custom colors
barplot(geary_vector,
        names.arg = marker_vector, cex.names = 0.4, las = 2,
        ylim = c(0,1),
        main = "Geary's C score", ylab = " Geary's C",
        col = bar_colors)  # Apply the custom color vector

# Add a horizontal line for the mean Geary's C
abline(h = geary_mean, col = "grey", lty = 2)


```

### SNR

Signal-to-noise ratio can be a good indicator of how strong the observed signal is for a given channel. This can be a function of antibody staining and marker abundance, among other factors.

It may be worth considering removing features that perform poorly in SNR.

```{r SNR barchart}

processed_snr <- computeSNR(processed_geary, method = "gaussian", update_correspondence = TRUE)
filtered_processed_snr <- processed_snr$CorrespondenceMatrix[!is.na(processed_snr$CorrespondenceMatrix$mz_location), ]

# arrange by snr
snr_barplot_df <- dplyr::arrange(filtered_processed_snr, snr)

# set value for recommended snr cutoff threshold, currently 10th percentile
threshold <- quantile(snr_barplot_df$snr, 0.1)

# plot
snr_plot <- ggplot(snr_barplot_df, aes(x = reorder(marker, snr), y = snr, fill = snr <= threshold)) +
  geom_bar(stat = "identity") +
  
  # Custom fill colors based on the SNR values
  scale_fill_manual(values = c("#18B596", "#b51837")) +
  
  # Add horizontal lines
  geom_hline(yintercept = threshold, linetype = "dashed", color = "black") +
  
  # Add text labels near the horizontal lines
  annotate("text", x = 1, y = threshold + 0.03, label = "Filter: 10th percentile", color = "black", hjust = 0, vjust = -0.5, size = 3) +

  
  # Customizations
  labs(title = "SNR-based Filtering", x = NULL, y = "SNR") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 5))

# Display the plot
print(snr_plot)

```
### Geary-SNR plot

Channels with a high Geary's C score and low SNR score can be confidently determined to be weak channels. It can be important to look at both of these metrics to avoid abberant removal of markers. Consider the case of a marker with a strong SNR, but with very uniform staining due to the ubiquity of the marker in the sample. Using only Geary's C for feature selection may result in identifying this marker as performing poorly.

```{r trashcan plot}

# fixed thresholds
geary_threshold <- 0.8
snr_threshold <- 2

# conditions for fill
geary_condition <- filtered_processed_snr$GearysC > geary_threshold
snr_condition <- filtered_processed_snr$snr < snr_threshold
filter <- geary_condition & snr_condition
filtered_processed_snr$fill_color <- ifelse(filter, "#b51837", "#18B596")

# Define the trashcan condition: SNR ≤ 2 and Geary's C ≥ 0.8
trashcan_zone <- filtered_processed_snr |>
  dplyr::filter(snr <= snr_threshold & GearysC >= geary_threshold)

# Create the trashcan plot with labels for markers in the trashcan zone
trashcan_plot <- ggplot() +
  theme_minimal() +
  ggtitle("Mean-SNR Plot: Trashcan plot") +
  # Shaded red rectangle (alpha = 0.2) above the vertical line and below the horizontal line
  geom_rect(aes(xmin = geary_threshold, xmax = 1, ymin = 0, ymax = snr_threshold), 
            fill = "red3", alpha = 0.5) +
  # Scatter plot for all markers
  geom_point(data = filtered_processed_snr, aes(x = GearysC, y = snr, fill = fill_color), 
             color = 'black', shape = 21, size = 4) +
  # Horizontal and vertical lines
  geom_hline(yintercept = snr_threshold, linetype = "solid", color = "black") +
  geom_vline(xintercept = geary_threshold, linetype = "solid", color = "black") +
  # Labels for points in the trashcan zone (markers with SNR ≤ 2 and Geary's C ≥ 0.8)
  ggrepel::geom_label_repel(data = trashcan_zone, aes(x = GearysC, y = snr, label = marker),
                   color = 'black',   # Border color
                   fill = 'white',    # Background color (inside the label)
                   box.padding = 0.2, 
                   point.padding = 0.2, 
                   max.overlaps = Inf, 
                   segment.color = 'black', # Line connecting the label to the point
                   size = 2, fontface = 'bold', label.size = 0.2) +  # label.size controls the border thickness
  # Use the fill color as defined in the data without further scaling
    #scale_x_log10() +
  scale_y_log10() +
  scale_fill_identity()

# Print the plot
print(trashcan_plot)

#ggplotly(trashcan_plot)

dput(sort(trashcan_zone$marker))


```

Here is the list of markers that perform poorly according to this plot, and are worth considering to remove from the feature list.

```{r print trashcan markers}

dput(sort(trashcan_zone$marker))

```


## Pixel-level QC

### TIC histogram thresholding

Thresholding on the total ion current of the sample can be potentially useful in terms of determining which pixels are off-tissue regions. Off-tissue pixels often have a lower TIC than tissue pixels.

```{r TIC histogram thresholding}

# Compute histogram
tic_targeted <- rowSums(processed$IntensityDF)
tic_targeted[tic_targeted == 0] <- NA
tic_targeted <- na.omit(tic_targeted)

threshold <- gutenTAG:::.otsu_thresholding(log10(tic_targeted))

# Compute the histogram data without plotting it
normalised_histogram <- hist(log10(1 + tic_targeted), 100, plot = FALSE)

# Plot the histogram with custom colors
bar_colors <- ifelse(normalised_histogram$mids <= threshold, "grey", "red3")

# Plot the histogram
plot(normalised_histogram, col = bar_colors, border = "black",
     main = "Normalised TIC histogram", cex.main = 2,
     xlab = "Total Ion Count (TIC)",
     xaxs = 'i', yaxs = 'i',
     cex.lab = 1.3)

# Add the threshold line
abline(v = threshold, lwd = 2, col = "grey", lty = 2)

```

