---
title: "preprocessing"
author: "John Abbey"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Install packages if necessary

```{r Install packages,message=FALSE}
#install.packages("Cardinal")
#nstall.packages("N2R")
#install.packages("EBImage")
#install.packages("imager")
```

Loading in Packages

```{r Packages,message=FALSE}
library(Cardinal)
library(N2R)
library(EBImage)
library(imager)
library(dplyr)
library(workflowr)
```

Load in functions

```{r Functions}
source("/mnt/msi_volume/Rscripts/maldi-processing/code/helper_functions.R")
```

Read in .imzML file

```{r Reading in data, message=FALSE}
Path_to_imzml_file = "/mnt/msi_volume/experiments/goodTonsil/menzha_20230323_tonwasopt_6_52034-20um-ag.imzML"
rawData = readMSIData(Path_to_imzml_file)
rawData
```

Optional get sample name step: splits path string into multiple strings and extract that which ends in .imzML

```{r Sample name, message=FALSE}
Sample_name <- sampleNameFinder(Path_to_imzml_file)
Sample_name
```

Read in panel

```{r Read in panel, echo=FALSE}
Path_to_peak_annotation = "/mnt/msi_volume/panels/30-plex-tonsil.csv"
peakAnnotation = read.delim(Path_to_peak_annotation, sep=",", header=TRUE, col.names = c("Name", "FeatureMass"))
peakAnnotation
```

Clean panel: orders by mass, cleans names, removes potentially problematic elements of string (e.g fullstop, spaces, etc.)

```{r Clean panel, echo=FALSE}
peakAnnotation <- cleanPanel(peakAnnotation)
```

Pre-process data

```{r Pre-processing}
preprocessed <- rawData %>% 
  Cardinal::normalize(method = "tic") %>%
  smoothSignal(method = "gaussian", plot=FALSE) %>%
  reduceBaseline(method="locmin") %>%
  #process(BPPARAM = MulticoreParam(workers = 4))
  process()

gc()
```

Pre-processed spectra visualisation. 

Plot by pixel number
```{r pressure, echo=FALSE}
plot(preprocessed, pixel = 1000)
```

Plot by pixel coordinate

```{r pressure, echo=FALSE}
plot(preprocessed, coord = c(100, 100))
```

Pre-processed image visualisation. Specify which mz you want to image.

```{r pressure, echo=FALSE}
image(preprocessed, mz = peakAnnotation$FeatureMass[1])
```

Write data

```{r Write pre-processed file , echo=FALSE}
writeMSIData(preprocessed, file = Sample_name, name=paste(Sample_name,"_pre-processed",sep = ""), folder = sample_path,
             mz.type = "32-bit float", intensity.type="32-bit float")
```


