---
title: "03_peak_detection"
author: "John Abbey"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

0. Load packages 
 
```{r Load packages, include=FALSE}
  suppressMessages(library(Cardinal))
  suppressMessages(library(fftw))
  suppressMessages(library(N2R))
  suppressMessages(library(Matrix))
  suppressMessages(library(igraph))
  suppressMessages(library(splus2R))
  suppressMessages(library(EBImage))
  suppressMessages(library(imager))
  suppressMessages(library(FactoMineR))
  suppressMessages(library(hexbin))
  suppressMessages(library(matrixStats))
  suppressMessages(library(mclust))
  suppressMessages(library(dplyr))
  suppressMessages(library(matter))
```

1. Get pre-processed intensity matrix

```{r Raw intensity}
raw_intensity <- iData(peakPre)
```

2. Get important dataset parameters.

```{r Parameter get}
range_peaks <- range(mz(peakPre))
n_features <- length(mz(peakPre))
n_pixels <- dim(peakPre)["Pixels"]
threshold_detection <- n_pixels*0.01
mz_vector <- as.data.frame(mz(peakPre))
location_pixels <- as.data.frame(pData(peakPre))[, c("x","y")]
```

3. Peak picking uses the cardinal peakPick function using median adaptive deviation, a SNR of 2 and a window with of 50mz.

```{r Peak picking}
list_peaks <- Cardinal::peakPick(peakPre, method = "mad", SNR = 2, window = 50) %>%
  process()
```

4. Obtain counts for each peak

```{r Obtain counts for each peak}

list_peaks <- peakData(list_peaks)[["mz"]]
n_peak_per_pixel <- unlist(lapply(list_peaks, FUN = length))

# unlist and order peaks
unlist_peaks <- unlist(list_peaks)
unlist_peaks_ordered <- unlist_peaks[order(unlist_peaks)]

# get peak frequencies
freq_peak_ordered <- table(factor(unlist_peaks_ordered, levels = mz_vector$mz))
freq_peak_ordered <- as.numeric(freq_peak_ordered)
```

5. Smooth over peak counts

```{r Smooth over peak counts }
# avg distance between features
mz_index_ratio <- (max(mz_vector$mz) - min(mz_vector$mz)) / n_features
# inverse of this ratio
sigma_smoothing_isotopic <- 1 / mz_index_ratio
# number of mz values centered around zero
list_values <- seq(-n_features / 2, n_features / 2, length.out = n_features)
# create the Gaussian for convolution
gaussian_vector <- exp(-list_values^2 / sigma_smoothing_isotopic^2)

# convolve peak frequencies with Gaussian vector
freq_peak_ordered_smooth <- stats::convolve(y = gaussian_vector, x = freq_peak_ordered, conj = TRUE)
# counts of rescaled ordered smooth peaks
freq_peak_ordered_smooth <- freq_peak_ordered_smooth[c((n_features / 2):n_features, 1:(n_features/2 - 1))]
```

6. Set count values below the threshold to zero

```{r Set count values below the threshold to zero}
freq_peak_smooth_thresholded <- freq_peak_ordered_smooth
freq_peak_smooth_thresholded[freq_peak_smooth_thresholded < threshold_detection] <- 0
```

7. Get metapeaks and their parameters

```{r  Get metapeaks and their parameters}
span_local_maxima <- 3 / mz_index_ratio
# round up to nearest odd value
local_maxima_peak_freq <- base::which(find_peaks(freq_peak_smooth_thresholded, span = span_local_maxima))
# flattened
local_maxima_peak_freq_reshaped <- matrix(rep(0, length(freq_peak_ordered)), ncol = 1)
local_maxima_peak_freq_reshaped[local_maxima_peak_freq] <- 1:length(local_maxima_peak_freq)

freq_peak_smooth_thresholded <- matrix(freq_peak_smooth_thresholded, ncol = 1)

# this is the metapeak selection step
propagation_selection <- propagate(freq_peak_smooth_thresholded, 
                                   seeds = local_maxima_peak_freq_reshaped, 
                                   mask = freq_peak_smooth_thresholded > threshold_detection)

metapeak_location <- aggregate.data.frame(data.frame(Location = mz_vector$mz, 
                                                     Freq = freq_peak_ordered),
                                         by = list(as.numeric(propagation_selection)), 
                                         FUN = function(x) {sum(x[[1]]*x[[2]]) / sum(x[[2]])})
metapeak_center <- c()
metapeak_max <- c()
metapeak_delimitation <- c()
for (k in 1:max(propagation_selection)) {
  # Looking for weighted mean
  location_centered_temp <- weighted.mean(x = mz_vector$mz[as.numeric(propagation_selection) == k],
                                         w = freq_peak_ordered[as.numeric(propagation_selection) == k] )
  # vector of centered peak locations
  metapeak_center <- c(metapeak_center, location_centered_temp)

  location_max_temp <- (mz_vector$mz[as.numeric(propagation_selection) == k])[which.max(freq_peak_ordered[as.numeric(propagation_selection) == k])]
  # vector of peak maximimum locations
  metapeak_max <- c(metapeak_max, location_max_temp)

  cumsum_freq <- (cumsum(freq_peak_ordered[as.numeric(propagation_selection) == k]) / sum(freq_peak_ordered[as.numeric(propagation_selection) == k]))

  beginning_peak <- max(which(cumsum_freq < 0.01))
  end_peak <- min(which(cumsum_freq > 0.99))
  if (is.infinite(beginning_peak)) {
    Beginning_peak = 1
  }
  # beginning and end of peak location
  metapeak_delimitation <- rbind(metapeak_delimitation, 
                                 c(mz_vector$mz[as.numeric(propagation_selection) == k][beginning_peak], 
                                   mz_vector$mz[as.numeric(propagation_selection) == k][end_peak]))
}
# width of each peak
metapeak_width <- metapeak_delimitation[, 2] - metapeak_delimitation[, 1]
# width of each peak on m/z scale
metapeak_width <- metapeak_width * mz_index_ratio
cat("...done ! \n")
```

8. Generate correspondence matrix

```{r Generate correspondence matrix}
mapping_meta <- crossKnn(mA = matrix(metapeak_max, ncol = 1),
                        mB= matrix(peakAnnotation_extended$Location, ncol = 1), k = 10, indexType = "L2", verbose = FALSE)
mz_threshold_association <- 1
# remove all mappings below the m/z distance association threshold
mapping_meta[mapping_meta > mz_threshold_association] <- 0
mapping_meta_cleaned <- apply(as.matrix(mapping_meta), MARGIN = 2, FUN = which_max_modified)

associated_marker <- c()
for (k in 1:length(metapeak_max)) {
  selected_mz_location <- range(mz_vector$mz[as.numeric(propagation_selection) == k])
  annotation_assigned <- peakAnnotation_extended$FeatureAnnotation[peakAnnotation_extended$Location > selected_mz_location[1] & peakAnnotation_extended$Location < selected_mz_location[2]]
  if (length(annotation_assigned) == 0){
    annotation_assigned = NA
  }
  if (length(annotation_assigned) > 1){
    annotation_assigned = paste(annotation_assigned, collapse="__OR__")
  }
  associated_marker <- c(associated_marker, annotation_assigned)
}

correspondence_matrix <- data.frame(mz_location = metapeak_max,
                                   Expected_mz_location = peakAnnotation_extended$Location[mapping_meta_cleaned],
                                   Annotation_peaks = peakAnnotation_extended$FeatureAnnotation[mapping_meta_cleaned])
rownames(peakAnnotation_extended) <- peakAnnotation_extended$FeatureAnnotation

```

9. Generate the Final Intensity DF

```{r Generate Final Intensity DF}
final_intensity <- c()
for (k in 1:nrow(Correspondence_matrix)) {
  # which mz locations are inside the peak (boolean)
  selected_mz_location = mz_vector$mz >= metapeak_delimitation[k,1] & mz_vector$mz <= metapeak_delimitation[k,2]
  # those mz values
  mz_vector$mz[which(selected_mz_location == TRUE)]
  if (sum(selected_mz_location) > 1) {
    intensity_temp = colSums(raw_intensity[selected_mz_location, ])
  }
  # if there is only one mz location, the intensity list is only the intensity values of the one mz location
  if (sum(selected_mz_location) == 1) {
    intensity_temp = (raw_intensity[selected_mz_location, ])
  }
  #intensity_temp = Correspondence_matrix$Annotation_peaks[k]
  final_intensity = cbind(final_intensity, intensity_temp)

}
cat("...done ! \n")

### DEV: GMM analysis on Final intensity matrix

# which of the metapeaks in the final intensity matrix are annotated peaks
final_intensity_targeted <- final_intensity[, !is.na(Correspondence_matrix$Annotation_peaks)]
colnames(final_intensity_targeted) <- correspondence_matrix$Annotation_peaks[!is.na(correspondence_matrix$Annotation_peaks)]
final_intensity_targeted <- as.data.frame(final_intensity_targeted)

```
